---
title: "Functional_Factors_Analysis"
author: "Zhongzhi"
date: "2026-03-16"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


### PERMANOVA analysis for beta diversity against environmental factors
```{r}
### To be modified Note
#Remove Sample NW046
#Select one of the three NW016 Sample.
library(dplyr)
library(tibble)
library(readr)
library(vegan)
library(ggplot2)
library(ggrepel)
library(compositions)
library(ape)

# Read in Normalized Pathway Abundance
path_abun_table <- file.path("../../data/Metagenomics/HUMAnN_func/Normalized_Tables/", "merged_pathabundance_relab.tsv")
path_abun <- read.table(path_abun_table, header = TRUE, sep = "\t", comment.char = "")


### Remove "Unmapped" And/OR "Unintegrated"
colnames(path_abun)[1]
colnames(path_abun)[1] <- "Pathway"
nrow(path_abun)
# # Remove both "Unmapped" and "Unintegrated"
# select_path_abun <- path_abun %>% filter(!Pathway %in% c("UNMAPPED", "UNINTEGRATED"))
# Only remove both "Unmapped"
select_path_abun <- path_abun %>% filter(!Pathway %in% c("UNMAPPED"))

### Remove pathways with taxonomic annotations
select_path_abun <- select_path_abun %>% filter (!grepl("\\|", Pathway))

# Total number of the pathways
nrow(select_path_abun)

# Remove not interested samples (without metadata and multiple sample for one individual)
ncol(select_path_abun)
select_path_abun <- select_path_abun %>% select(-NW046_5_all_merged_Abundance)
select_path_abun <- select_path_abun %>% select(-NW016_2_2_all_merged_Abundance) %>% select(-NW016_5_all_merged_Abundance)

ncol(select_path_abun)

# Filter on the prevalence (and abundance) of the Pathways
# Prevalence > 10%, Abundance > 0.1%/0.01%
# Convert into matrix：rows=Pathway, cols=Sample
mat_path <- select_path_abun %>%
  column_to_rownames("Pathway") %>%
  as.matrix()
storage.mode(mat_path) <- "numeric"

# Matrix: samples x pathways
mat_samp <- t(mat_path)

mat_rel <- mat_samp

# Prevalence (and Abundance) filtering
prev <- colMeans(mat_rel > 0, na.rm = TRUE)              
mean_rel <- colMeans(mat_rel, na.rm = TRUE)
prev_cut <- 0.10

keep <- (prev >= prev_cut)

mat_rel_filt <- mat_rel[, keep, drop = FALSE]

# The number of left pathways
ncol(mat_rel_filt)

### Calculate the Bray-Curtis distance
dist_bc <- vegdist(mat_rel_filt, method = "bray")


### Calculate the Aitchison distance
# add a small pseudocount
pseudo <- 1e-6
mat_rel_pc <- mat_rel_filt + pseudo
# re-close to composition (optional but recommended)
mat_rel_pc <- mat_rel_pc / rowSums(mat_rel_pc)
# CLR transform
mat_clr <- clr(mat_rel_pc)
# Aitchison distance = Euclidean distance on CLR data
dist_aitchison <- dist(mat_clr, method = "euclidean")



###### Analysis based on Bray-Curtis distance
### PCoA Plot
# pcoa <- cmdscale(dist_bc, k = 2, eig = TRUE)
# pcoa_df <- as.data.frame(pcoa$points)
# colnames(pcoa_df) <- c("PCoA1", "PCoA2")
# pcoa_df$SampleID_in_table <- rownames(pcoa_df)
# 
# # Explanation (Percentage)
# var_exp <- pcoa$eig / sum(pcoa$eig)
# pcoa1_lab <- paste0("PCoA1 (", round(100 * var_exp[1], 1), "%)")
# pcoa2_lab <- paste0("PCoA2 (", round(100 * var_exp[2], 1), "%)")


###### Analysis based on Aitchison distance
pcoa_res <- ape::pcoa(dist_aitchison)
pcoa_df <- as.data.frame(pcoa_res$vectors[, 1:2])
colnames(pcoa_df) <- c("PCoA1", "PCoA2")
pcoa_df$SampleID_in_table <- rownames(pcoa_df)

# Calculate percentage of explained variation using positive eigenvalues only
pcoa1_lab <- paste0("PCoA1 (", round(pcoa_res$values$Relative_eig[1] * 100, 1), "%)")
pcoa2_lab <- paste0("PCoA2 (", round(pcoa_res$values$Relative_eig[2] * 100, 1), "%)")

### Read in metadata
meta_file <- file.path("../../data/metadata", "selected_northwood_map.txt")
metadata <- read_tsv(meta_file, col_types = cols())

# Extract individual ID from metadata ("#SampleID", "NW001-1")
metadata <- metadata %>%
  mutate(Sample_prefix = sub("-.*$", "", `#SampleID`))
head(metadata)

# Extract individual ID from pathway table ("NW001_3_all_merged_Abundance")
pcoa_df <- pcoa_df %>%
  mutate(Sample_prefix = sub("_.*$", "", SampleID_in_table)) %>%
  left_join(metadata, by = "Sample_prefix")



# ### Filtering on PCoA df
pcoa_df <- pcoa_df %>%
  mutate(Update_SampleID = sub("_all_merged.*$", "", SampleID_in_table))
# head(pcoa_df)
# nrow(pcoa_df)
# filtered_pcoa_df <- pcoa_df %>% filter(`#SampleID` != "NW046-1")
# nrow(filtered_pcoa_df)
# filtered_pcoa_df %>% filter(`#SampleID` == "NW016-1")

filtered_pcoa_df <- pcoa_df



### PCoA Plot Colored with Sex (Colored based on FIScore)
p_pcoa <- ggplot(filtered_pcoa_df, aes(x = PCoA1, y = PCoA2, color = FIScore)) +
  geom_point(size = 2, alpha = 0.8) +
  geom_text_repel(aes(label = Update_SampleID),
                  size = 3,
                  max.overlaps = 20) +
  scale_color_viridis_c(option = "plasma",
                        limits = c(0.3, 0.8),
                        oob = scales::squish) +
  theme_bw(base_size = 12) +
  labs(
    #title = "PCoA of pathway abundance (Bray-Curtis)",
    title = "PCoA of pathway abundance (Aitchison)",
    x = pcoa1_lab,
    y = pcoa2_lab
  ) +
  theme(legend.position = "right",
        plot.title = element_text(hjust = 0.5, face = "bold"))

p_pcoa

# figure_file <- file.path("../../tmp_figures/PCoA_Plots", "Pathways_PCoA_Plot_labeled_rm_samp.jpg")
# ggsave(filename = figure_file, plot = p_pcoa, width = 8, height = 6, dpi = 300)





###### PERMANOVA Analyisis for All Other Factors
### Analysis based on Aitchison distance instead of Bray-Curtis distance
dist_bc <- dist_aitchison

meta <- filtered_pcoa_df %>%
  distinct(SampleID = SampleID_in_table, Sex, FIRange, FICat, Age, TimeNW, FIScore, Location)

meta <- meta[match(labels(dist_bc), meta$SampleID), ]
stopifnot(all(meta$SampleID == labels(dist_bc)))

head(meta)
tail(meta)


set.seed(1)
# Test with FIScore
adonis_fiscore <- adonis2(dist_bc ~ FIScore, data = meta, permutations = 9999)
adonis_fiscore
# Test with Age
adonis_age <- adonis2(dist_bc ~ Age, data = meta, permutations = 9999)
adonis_age
# Test with TimeNW
adonis_timenw <- adonis2(dist_bc ~ TimeNW, data = meta, permutations = 9999)
adonis_timenw 
# Test with Location
adonis_location <- adonis2(dist_bc ~ Location, data = meta, permutations = 9999)
adonis_location
# Test with Sex
adonis_sex <- adonis2(dist_bc ~ Sex, data = meta, permutations = 9999)
adonis_sex
# Test with FICat
adonis_ficat <- adonis2(dist_bc ~ FICat, data = meta, permutations = 9999)
adonis_ficat


###### Multiple variable test
adonis_multi <- adonis2(
  dist_bc ~ Sex + Age + TimeNW + FIScore + Location,
  data = meta,
  permutations = 9999,
  by = "margin"
)
adonis_multi

# Output Results
res_multi <- as.data.frame(adonis_multi)
res_multi$Variable <- rownames(res_multi)
rownames(res_multi) <- NULL

res_multi_out <- res_multi %>%
  select(Variable, Df, R2, F, `Pr(>F)`)

res_multi_out

res_multi_file_name <- file.path("../../tmp_figures/PCoA_Plots", "pathway_adonis_multi_summary.tsv")
write.table(res_multi_out, res_multi_file_name,
            sep = "\t", quote = FALSE, row.names = FALSE)



# Dispersion Checking
disp_sex <- betadisper(dist_bc, meta$Sex)
anova(disp_sex)
permutest(disp_sex, permutations = 9999)

disp_ficat <- betadisper(dist_bc, meta$FICat)
anova(disp_ficat)
permutest(disp_ficat, permutations = 9999)

disp_location <- betadisper(dist_bc, meta$Location)
anova(disp_location)
permutest(disp_location, permutations = 9999)


### PCOA Plots for factor variables (FICat)
filtered_pcoa_df <- filtered_pcoa_df %>%
  mutate(FICat = factor(FICat, levels = sort(unique(FICat))))

r2_ficat <- adonis_ficat$R2[1]
p_ficat  <- adonis_ficat$`Pr(>F)`[1]

pcoa_ficat <- ggplot(filtered_pcoa_df, aes(x = PCoA1, y = PCoA2, color = FICat)) +
  geom_point(size = 2, alpha = 0.8) +
  scale_color_viridis_d(option = "plasma", end = 0.95) +
  theme_bw(base_size = 12) +
  labs(
    title = "PCoA of pathway abundance (Bray-Curtis)",
    subtitle = sprintf("PERMANOVA (FICat): R² = %.3f, p = %.4g", r2_ficat, p_ficat),
    x = pcoa1_lab,
    y = pcoa2_lab
  ) +
  stat_ellipse(aes(group = FICat), type = "t", level = 0.95,
               linewidth = 0.8, show.legend = FALSE) +
  theme(
    legend.position = "right",
    plot.title = element_text(hjust = 0.5, face = "bold")
  )

pcoa_ficat

figure_file <- file.path("../../tmp_figures/PCoA_Plots", "Pathways_PCoA_Plot_FICat_grouped_color.jpg")
ggsave(filename = figure_file, plot = pcoa_ficat, width = 8, height = 6, dpi = 300)


### PCOA Plots for factor variables (Location)
r2_location <- adonis_location$R2[1]
p_location  <- adonis_location$`Pr(>F)`[1]

pcoa_location <- ggplot(filtered_pcoa_df, aes(x = PCoA1, y = PCoA2, color = Location)) +
  geom_point(size = 2, alpha = 0.8) +
  scale_color_viridis_d(option = "plasma", end = 0.95) +
  theme_bw(base_size = 12) +
  labs(
    title = "PCoA of pathway abundance (Bray-Curtis)",
    subtitle = sprintf("PERMANOVA (Location): R² = %.3f, p = %.4g", r2_location, p_location),
    x = pcoa1_lab,
    y = pcoa2_lab
  ) +
  stat_ellipse(aes(group = Location), type = "t", level = 0.95,
               linewidth = 0.8, show.legend = FALSE) +
  theme(
    legend.position = "right",
    plot.title = element_text(hjust = 0.5, face = "bold")
  )

pcoa_location

figure_file <- file.path("../../tmp_figures/PCoA_Plots", "Pathways_PCoA_Plot_Location_grouped_color.jpg")
ggsave(filename = figure_file, plot = pcoa_location, width = 8, height = 6, dpi = 300)


### PCOA Plots for factor variables (Sex)
r2_sex <- adonis_sex$R2[1]
p_sex  <- adonis_sex$`Pr(>F)`[1]

pcoa_sex <- ggplot(filtered_pcoa_df, aes(x = PCoA1, y = PCoA2, color = Sex)) +
  geom_point(size = 2, alpha = 0.8) +
  scale_color_viridis_d(option = "plasma", end = 0.95) +
  theme_bw(base_size = 12) +
  labs(
    title = "PCoA of pathway abundance (Bray-Curtis)",
    subtitle = sprintf("PERMANOVA (Sex): R² = %.3f, p = %.4g", r2_sex, p_sex),
    x = pcoa1_lab,
    y = pcoa2_lab
  ) +
  stat_ellipse(aes(group = Sex), type = "t", level = 0.95,
               linewidth = 0.8, show.legend = FALSE) +
  theme(
    legend.position = "right",
    plot.title = element_text(hjust = 0.5, face = "bold")
  )

pcoa_sex

figure_file <- file.path("../../tmp_figures/PCoA_Plots", "Pathways_PCoA_Plot_Sex_grouped_color.jpg")
ggsave(filename = figure_file, plot = pcoa_sex, width = 8, height = 6, dpi = 300)


### PCoA Plots for continuous variables (FIScore)
r2_fiscore <- adonis_fiscore$R2[1]
p_fiscore  <- adonis_fiscore$`Pr(>F)`[1]

p_fiscore <- ggplot(filtered_pcoa_df, aes(x = PCoA1, y = PCoA2, color = FIScore)) +
  geom_point(size = 2, alpha = 0.8) +
  scale_color_viridis_c(option = "plasma", end = 0.95) +
  theme_bw(base_size = 12) +
  labs(
    title = "PCoA of pathway abundance (Bray-Curtis)",
    subtitle = sprintf("PERMANOVA (FIScore): R² = %.3f, p = %.4g", r2_fiscore, p_fiscore),
    x = pcoa1_lab,
    y = pcoa2_lab
  ) +
  theme(
    legend.position = "right",
    plot.title = element_text(hjust = 0.5, face = "bold")
  )

p_fiscore

figure_file <- file.path("../../tmp_figures/PCoA_Plots", "Pathways_PCoA_Plot_FIScore_color.jpg")
ggsave(filename = figure_file, plot = p_fiscore, width = 8, height = 6, dpi = 300)



### PCoA Plots for continuous variables (Age)
r2_age <- adonis_age$R2[1]
p_age  <- adonis_age$`Pr(>F)`[1]

pcoa_age <- ggplot(filtered_pcoa_df, aes(x = PCoA1, y = PCoA2, color = Age)) +
  geom_point(size = 2, alpha = 0.8) +
  scale_color_viridis_c(option = "plasma", end = 0.95) +
  theme_bw(base_size = 12) +
  labs(
    title = "PCoA of pathway abundance (Bray-Curtis)",
    subtitle = sprintf("PERMANOVA (Age): R² = %.3f, p = %.4g", r2_age, p_age),
    x = pcoa1_lab,
    y = pcoa2_lab
  ) +
  theme(
    legend.position = "right",
    plot.title = element_text(hjust = 0.5, face = "bold")
  )

pcoa_age


figure_file <- file.path("../../tmp_figures/PCoA_Plots", "Pathways_PCoA_Plot_Age_color.jpg")
ggsave(filename = figure_file, plot = pcoa_age, width = 8, height = 6, dpi = 300)




### PCoA Plots for continuous variables (TimeNW)
r2_timenw <- adonis_timenw$R2[1]
p_timenw   <- adonis_timenw$`Pr(>F)`[1]

pcoa_timenw <- ggplot(filtered_pcoa_df, aes(x = PCoA1, y = PCoA2, color = TimeNW)) +
  geom_point(size = 2, alpha = 0.8) +
  scale_color_viridis_c(option = "plasma", end = 0.95) +
  theme_bw(base_size = 12) +
  labs(
    title = "PCoA of pathway abundance (Bray-Curtis)",
    subtitle = sprintf("PERMANOVA (TimeNW): R² = %.3f, p = %.4g", r2_timenw, p_timenw),
    x = pcoa1_lab,
    y = pcoa2_lab
  ) +
  theme(
    legend.position = "right",
    plot.title = element_text(hjust = 0.5, face = "bold")
  )
pcoa_timenw

figure_file <- file.path("../../tmp_figures/PCoA_Plots", "Pathways_PCoA_Plot_TimeNW_color.jpg")
ggsave(filename = figure_file, plot = pcoa_timenw, width = 8, height = 6, dpi = 300)


```



### Study Pathway Relative Abundance against environmental factors
```{r}
library(dplyr)
library(readr)
library(tidyr)
library(stringr)
library(ggplot2)
library(tibble)
library(scales)
library(pheatmap)

### Readin Humann Pathway Abundance Table
infile <- file.path("../../data/Metagenomics/HUMAnN_func/Normalized_Tables", "merged_pathabundance_relab.tsv")
#outfile <- "summed_path_abundance.out"
corr_file <- file.path("../../tmp_figures", "pathway_metadata_correlations.tsv")

path_df <- read_tsv(infile) %>% rename(Pathway = `# Pathway`)


### New and more direct method to extract the total abundance of each pathway
head(path_df)
nrow(path_df)

path_merged <- path_df %>%
  filter (!grepl("\\|", Pathway))
head(path_merged)
nrow(path_merged)


# Filter pathway abundance data for pathways exists in at least 1/3 of all MG samples
sample_cols <- names(path_merged)[-1]
n_samples <- length(sample_cols)
min_present <- ceiling(n_samples / 3)
min_present

path_merged <- path_merged %>%
  filter(
    rowSums(across(all_of(sample_cols), ~ .x > 0), na.rm = TRUE) >= min_present
  )
nrow(path_merged)

### Output Results
#write.table(path_merged, outfile, row.names = FALSE, sep = "\t", quote = FALSE)


### Read in Samples Meta Table
meta_file <- "../../data/Metadata/selected_northwood_map.txt"
metadata <- read_tsv(meta_file, col_types = cols())
# Extract individual names: NW001, NW002 ...
metadata <- metadata %>%
  mutate(Sample_prefix = sub("-.*$", "", `#SampleID`))
head(metadata)
nrow(metadata)

path_long <- path_merged %>%
  pivot_longer(
    cols      = -Pathway,
    names_to  = "Sample_col",
    values_to = "Abundance"
  ) %>%
  # Extract individual names: NW001, NW002 ...
  mutate(Sample_prefix = sub("_.*$", "", Sample_col))

head(path_long)
nrow(path_long)
#table(path_long$Pathway)
#table(path_long$Sample_col)


### Remove duplicate NW016 samples in the MG pathway abundance table
selected_path_long <- path_long %>% filter(Sample_col != "NW016_2_2_all_merged_Abundance") %>% filter(Sample_col != "NW016_5_all_merged_Abundance")
nrow(selected_path_long)

### Remove the NW046 sample which does not have Metadata
selected_path_long <- selected_path_long %>% filter(Sample_col != "NW046_5_all_merged_Abundance")
nrow(selected_path_long)



### Add metadata to the MG pathway abundance table
path_meta <- selected_path_long %>% left_join(metadata, by = "Sample_prefix")

# Modification to variables
path_meta2 <- path_meta %>%
  mutate(
    Sex_num = case_when(
      Sex == "Male"   ~ 1,
      Sex == "Female" ~ 0,
      TRUE            ~ NA_real_
    )
  ) %>%
  mutate(
    Location_num = case_when(
      Location == "Center"   ~ 1,
      Location == "Manor" ~ 0,
      TRUE            ~ NA_real_
    )
  )  

vars <- c("Age", "TimeNW", "FIScore", "Sex_num", "FICat", "Location_num")
binary_vars <- c("Sex", "Location", "FICat")



###### Calculate pathway differences across binary factors
binary_list <- lapply(binary_vars, function(v) {
  path_meta %>%
    filter(!is.na(.data[[v]]), !is.na(Abundance)) %>%
    group_by(Pathway) %>%
    summarise(
      n = n(),
      n_group = n_distinct(.data[[v]]),
      stat = if (
        n_group[1] == 2
      ) {
        suppressWarnings(wilcox.test(Abundance ~ .data[[v]])$statistic)
      } else {
        NA_real_
      },
      p = if (
        n_group[1] == 2
      ) {
        suppressWarnings(wilcox.test(Abundance ~ .data[[v]])$p.value)
      } else {
        NA_real_
      },
      .groups = "drop"
    ) %>%
    mutate(variable = v, test = "wilcox") %>%
    select(Pathway, variable, test, n, n_group, stat, p)
})

test_table_binary <- bind_rows(binary_list)
test_table_binary <- test_table_binary %>%
  group_by(variable) %>%
  mutate(p_adj = p.adjust(p, method = "fdr")) %>%
  ungroup()


head(test_table_binary)

table(test_table_binary$variable)
filtered_test_table_binary <- test_table_binary %>% filter(p_adj < 0.05)
nrow(filtered_test_table_binary)
# filtered_test_table_binary <- test_table_binary %>% filter(p < 0.01)
# nrow(filtered_test_table_binary)

table_file <- file.path("../../tmp_figures", "pathway_metadata_binary_test.tsv")
write_tsv(test_table_binary, table_file)



###### Heatmap plot for pathways with significant changes across locations
loc_sig_pathways <- filtered_test_table_binary %>%
  filter(variable == "Location") %>%
  pull(Pathway) %>%
  unique()

# Extract pathway abundance matrix
heatmap_df <- selected_path_long %>%
  filter(Pathway %in% loc_sig_pathways)

heatmap_df <- heatmap_df %>%
  select(Pathway, Sample_col, Abundance) %>%
  pivot_wider(
    names_from = Sample_col,
    values_from = Abundance
  )

heatmap_mat <- heatmap_df %>%
  column_to_rownames("Pathway") %>%
  as.matrix()

storage.mode(heatmap_mat) <- "numeric"

# Prepare sample annotation using metadata
sample_anno <- data.frame(
  SampleID_in_table = colnames(heatmap_mat),
  Sample_prefix = sub("_.*$", "", colnames(heatmap_mat))
) %>%
  left_join(metadata %>% select(Sample_prefix, Location), by = "Sample_prefix")

annotation_col <- sample_anno %>%
  select(Location)
rownames(annotation_col) <- sample_anno$SampleID_in_table
annotation_col <- annotation_col[colnames(heatmap_mat), , drop = FALSE]

# Simplify column names in the heatmap matrix
new_colnames <- sub("_all_merged_Abundance.*$", "", colnames(heatmap_mat))
colnames(heatmap_mat) <- new_colnames
rownames(annotation_col) <- new_colnames

# Optional annotation colors
ann_colors <- list(
  Location = c(
    Center = "#1f78b4",
    Manor = "#33a02c"
  )
)


### Draw and save heatmap
figure_file <- file.path("../../tmp_figures/", "Location_Pathways_heatmap.jpg")
jpeg(filename = figure_file, width = 8, height = 6, units = "in", res = 300)

pheatmap(
  heatmap_mat,
  scale = "row",
  annotation_col = annotation_col,
  annotation_colors = ann_colors,
  cluster_rows = TRUE,
  cluster_cols = TRUE,
  show_rownames = TRUE,
  show_colnames = TRUE,
  fontsize_row = 8,
  fontsize_col = 7,
  main = "Significant pathways associated with Location\n(Row Z-score)"
)
#print(loc_path_heatmap)

dev.off()



###### Calculate correlations between pathway abundance and individual factors
corr_list <- lapply(vars, function(v) {
  # Rename Sex_num, Location_num
  var_label <- dplyr::case_when(
    v == "Sex_num"      ~ "Sex",
    v == "Location_num" ~ "Location",
    TRUE ~ v
  )
  
  path_meta2 %>%
    group_by(Pathway) %>%
    summarise(
      n = sum(!is.na(Abundance) & !is.na(.data[[v]])),
      cor = ifelse(
        n >= 3,
        suppressWarnings(cor(Abundance, .data[[v]],
                             method = "spearman",
                             use = "complete.obs")),
        NA_real_
      ),
      p = ifelse(
        n >= 3,
        suppressWarnings(cor.test(Abundance, .data[[v]],
                                  method = "spearman",
                                  exact = FALSE)$p.value),
        NA_real_
      ),
      .groups = "drop"
    ) %>%
    mutate(variable = var_label) %>%
    select(Pathway, variable, n, cor, p)
})

corr_table <- bind_rows(corr_list)
head(corr_table)
table(corr_table$variable)

### FDR adjust
corr_table <- corr_table %>%
  group_by(variable) %>%
  mutate(p_adj = p.adjust(p, method = "fdr")) %>%
  ungroup()

# ### Output correlation results
# write_tsv(corr_table, corr_file)


### Quick Sta
#corr_table <- read.table(corr_file, header = TRUE, sep = "\t")
#p-adj filter
nrow(corr_table %>% filter(p_adj < 0.05))
p_adj_filter_corr_table <- corr_table %>% filter(p_adj < 0.05)
table(p_adj_filter_corr_table$variable)

# #p value filter
# nrow(corr_table %>% filter(p < 0.01))
# p_filter_corr_table <- corr_table %>% filter(p < 0.01)
# table(p_filter_corr_table$variable)







###### Scatter Plots for pathways
### Plots for age-associated pathways
# Select significantly age-associated pathways in corr_table
paths_of_interest <- corr_table %>%
  filter(p_adj < 0.05, variable == "Age") %>%
  pull(Pathway) %>%
  unique()

paths_of_interest

plot_df <- path_meta2 %>%
  filter(Pathway %in% paths_of_interest)

# Log10 transfer
make_scatter <- function(df, pw_name) {
  safe_name <- pw_name %>%
    str_replace(":.*$", "") 

  ggplot(df, aes(x = Age, y = Abundance)) +  
    geom_point(size = 2, alpha = 0.5) +
    geom_smooth(method = "lm", se = TRUE, color = "red") +
    scale_y_continuous(labels = label_scientific()) +
    theme_bw(base_size = 12) +
    theme(
      panel.grid = element_blank(),
      plot.title = element_text(face = "bold", hjust = 0.5),
      axis.title = element_text(face = "bold"),
      axis.text = element_text(color = "black")
    ) +
    labs(
      title = safe_name,
      x = "Age",
      y = "Pathway relative abundance"
    )
}



for (pw in paths_of_interest) {
  
  df_pw <- plot_df %>% filter(Pathway == pw)

  # Generate shorter file name
  safe_name <- pw %>%
    str_replace(":.*$", "") 

  p <- make_scatter(df_pw, pw)
  
  figure_file <- file.path("../../tmp_figures", paste0("Age_vs_", safe_name, ".jpg"))

  ggsave(
    filename = figure_file,
    plot     = p,
    width    = 5,
    height   = 4,
    dpi      = 300
  )
}

```



### Analysis with MaAsLin3
```{r}
library(dplyr)
library(maaslin3)

### Read in HUMANn pathway abundance table
data_table_name <- file.path("../../data/Metagenomics/HUMAnN_func/Normalized_Tables", "merged_pathabundance_relab.tsv")

# data_table_name <- file.path("../../data/Metagenomics/HUMAnN_func/", "nursing_house_humman_cov30_path_abundance.out")

org_data_table <- read.csv(data_table_name, sep = '\t', row.names = 1)
nrow(org_data_table)
clean_data_table <- org_data_table[!grepl("\\|", rownames(org_data_table)), ]
nrow(clean_data_table)

# Remove "UNMAPPED" only
clean_data_table <- clean_data_table[!rownames(clean_data_table) %in% c("UNMAPPED"), ]

# # Remove "UNINTEGRATED" only
# clean_data_table <- clean_data_table[!rownames(clean_data_table) %in% c("UNINTEGRATED"), ]

# # Remove "UNMAPPED" and "UNINTEGRATED"
# clean_data_table <- clean_data_table[!rownames(clean_data_table) %in% c("UNMAPPED", "UNINTEGRATED"), ]

# # Update pathway names for better visualization
# rownames(clean_data_table) <- sub(":.*", "", rownames(clean_data_table))

nrow(clean_data_table)
data_table <- t(clean_data_table)
ncol(data_table)


### Filtering on Pathway Prevalence
# Calculate pathway prevalence across samples
prev <- colMeans(data_table > 0, na.rm = TRUE)

# Keep pathways with prevalence > 0.1
prev_cut <- 0.10
keep_pathways <- prev > prev_cut
length(keep_pathways)
length(unique(keep_pathways))

# Filter pathways
data_table <- data_table[, keep_pathways, drop = FALSE]

# Check number of retained pathways
ncol(data_table)


# Remove Duplicate Samples and Sample without metadata
data_table <- data_table[!rownames(data_table) %in% c("NW016_2_2_all_merged_Abundance", "NW016_5_all_merged_Abundance", "NW046_5_all_merged_Abundance"), ]
data_table[1:5, 1:5]
nrow(data_table)



# ### Read in HUMANn gene family abundance table
# data_table_name <- file.path("../../data/Metagenomics/HUMAnN_func", "nursing_house_humman_cov30_gene_families_anno.out")
# org_data_table <- read.csv(data_table_name, sep = '\t', row.names = 1, check.names = FALSE)
# nrow(org_data_table)
# clean_data_table <- org_data_table[!grepl("\\|", rownames(org_data_table)), ]
# nrow(clean_data_table)
# data_table <- t(clean_data_table)
# 
# # Remove Duplicate Samples and Sample without metadata
# data_table <- data_table[!rownames(data_table) %in% c("NW016_2_2_all_merged_Abundance-RPKs", "NW016_5_all_merged_Abundance-RPKs", "NW046_5_all_merged_Abundance-RPKs"), ]
# data_table[1:5, 1:5]
# nrow(data_table)
# ncol(data_table)
# # Remove column(gene family) with non-zero values in <10% rows(samples)
# nonzero_counts <- colSums(data_table != 0)
# threshold <- ceiling(0.8 * nrow(data_table))
# data_table<- data_table[, nonzero_counts >= threshold]
# ncol(data_table)



# ### Read in MetaPhlAn taxa abundance table
# data_table_name <- file.path("../../data/Metagenomics/MetaPhlAn_taxa", "merged_abundance_table.txt")
# org_data_table <- read.csv(data_table_name, sep = '\t', row.names = 1, comment.char = "#")
# nrow(org_data_table)
# 
# # Select rows with species level annotation
# clean_data_table <- org_data_table[
#   grepl("\\|s__", rownames(org_data_table)) &
#   !grepl("\\|t__", rownames(org_data_table)),
# ]
# 
# # # Select rows with genus level annotation
# # clean_data_table <- org_data_table[
# #   grepl("\\|g__", rownames(org_data_table)) &
# #   !grepl("\\|s__", rownames(org_data_table)),
# # ]
# 
# # Select rows with family level annotation
# clean_data_table <- org_data_table[
#   grepl("\\|f__", rownames(org_data_table)) &
#   !grepl("\\|g__", rownames(org_data_table)),
# ]
# 
# # # Select rows with phylum level annotation
# # clean_data_table <- org_data_table[
# #   grepl("\\|p__", rownames(org_data_table)) &
# #   !grepl("\\|c__", rownames(org_data_table)),
# # ]
# 
# nrow(clean_data_table)
# data_table <- t(clean_data_table)
# 
# # Remove Duplicate Samples and Sample without metadata
# data_table <- data_table[!rownames(data_table) %in% c("NW016_2_2_all", "NW016_5_all", "NW046_5_all"), ]
# # colnames(data_table) <- sub(".*\\|s__", "", colnames(data_table))
# # # colnames(data_table) <- sub(".*\\|g__", "", colnames(data_table))
# colnames(data_table) <- sub(".*\\|f__", "", colnames(data_table))
# # # colnames(data_table) <- sub(".*\\|p__", "", colnames(data_table))
# 
# 
# data_table[1:5, 1:5]
# nrow(data_table)
# ncol(data_table)


### Read in metadata table
metadata_name <- file.path("../../data/Metadata", "selected_northwood_map.txt")
org_metadata <- read.csv(metadata_name, sep = '\t', row.names = 1)
# Remove sample without metadata
metadata <- org_metadata[!rownames(org_metadata) %in% c("NW046-1"), ]
head(metadata)
nrow(metadata)

# Matching ID names
# 1. Extract Individual ID
data_id <- sub("_.*", "", rownames(data_table))
meta_id <- sub("-.*", "", rownames(metadata))
# 2. Matching
match_idx <- match(meta_id, data_id)
# 3. Remove unmatched ID
keep <- !is.na(match_idx)
metadata_matched <- metadata[keep, ]
match_idx <- match_idx[keep]
# 4. Replace/Update rownames of the metadata
rownames(metadata_matched) <- rownames(data_table)[match_idx]
metadata <- metadata_matched
head(metadata)
# Rename the FIScore column into FI
metadata <- metadata %>%
  rename(FI = FIScore)
metadata <- metadata %>%
  mutate(
    Location = recode(Location, "Center" = "Centre")
  )


### Factor the categorical variables in metadata
metadata$Sex <- 
  factor(metadata$Sex, levels = c('Male', 'Female'))

# metadata$Location <- 
#   factor(metadata$Location, levels = c('Manor', 'Center'))

metadata$Location <- 
  factor(metadata$Location, levels = c('Manor', 'Centre'))

metadata$FICat <- 
  factor(metadata$FICat, levels = c('3', '4'))

# Check dataframes
data_table[1:5, 1:5]
metadata[1:5, ]
tail(metadata)




# ### Running MaAsLin3
# set.seed(1)
# fit_out <- maaslin3(input_data = data_table,
#                     input_metadata = metadata,
#                     output = 'MaAsLin3_pathway_rm_unmap_test_output',
#                     formula = '~ Sex + Age +
#                         TimeNW + FIScore + Location',
#                     normalization = 'TSS',
#                     transform = 'LOG',
#                     coef_plot_vars = c("Age", "FIScore"),
#                     heatmap_vars = c("TimeNW", "Location Center", "Sex Female"),
#                     augment = TRUE,
#                     standardize = TRUE,
#                     max_significance = 0.1,
#                     median_comparison_abundance = TRUE,
#                     median_comparison_prevalence = FALSE,
#                     max_pngs = 250,
#                     cores = 4)


### Running MaAsLin3 (Updated)
set.seed(1)
fit_out <- maaslin3(input_data = data_table,
                    input_metadata = metadata,
                    output = 'MaAsLin3_pathway_rm_unmap_top15_output',
                    #output = 'MaAsLin3_MG_family_update_output',
                    formula = '~ Sex + Age +
                        TimeNW + FI + Location',
                    summary_plot_first_n = 15,
                    normalization = 'TSS',
                    transform = 'LOG',
                    coef_plot_vars = c("Age", "Location Centre"),
                    heatmap_vars = c("FI", "TimeNW", "Sex Female"),
                    augment = TRUE,
                    standardize = TRUE,
                    #max_significance = 0.05,
                    median_comparison_abundance = TRUE,
                    median_comparison_prevalence = FALSE,
                    max_pngs = 250,
                    cores = 4)

```



### Refined plots of associations between pathway abundance and age (Based on MaAsLin3 results)
```{r}
library(dplyr)
library(tidyr)
library(readr)
library(stringr)
library(ggplot2)
library(ggrepel)
library(tibble)

### Read in MaAsLin3 significant table
# maaslin_file <- file.path("MaAsLin3_pathway_rm_unmap_test_output", "significant_results.tsv")
maaslin_file <- file.path("MaAsLin3_pathway_rm_unmap_top35_output", "significant_results.tsv")
maaslin_table <- read_tsv(maaslin_file, col_types = cols())

# Check column names if needed
# colnames(maaslin_table)

selected_maaslin_table <- maaslin_table %>%
  #filter(metadata == "Age", qval_joint < 0.01)
  filter(metadata == "Age", qval_joint < 0.1)

selected_pathway <- unique(selected_maaslin_table$feature)
selected_pathway


### Read in sample metadata
metadata_file <- file.path("../../data/Metadata", "selected_northwood_map.txt")
metadata_table <- read_tsv(metadata_file, col_types = cols())

metadata_table <- metadata_table %>%
  rename(Sample = `#SampleID`) %>%
  select(Sample, Age, Sex, TimeNW, FIScore, Location)
head(metadata_table)


### Read in pathway abundance data
org_path_abun_file <- file.path(
  "../../data/Metagenomics/HUMAnN_func/Normalized_Tables",
  "merged_pathabundance_relab.tsv"
)

org_path_abun_table <- read.table(
  org_path_abun_file,
  header = TRUE,
  sep = "\t",
  comment.char = "",
  quote = "",
  check.names = FALSE
)

colnames(org_path_abun_table)[1] <- "Pathway"

### Remove stratified rows and unwanted rows
path_abun_merged <- org_path_abun_table %>%
  filter(!grepl("\\|", Pathway)) %>%
  filter(!Pathway %in% c("UNMAPPED"))

head(path_abun_merged)
nrow(path_abun_merged)


### Re-close abundance data after removing "UNMAPPED"
mat_path <- path_abun_merged %>%
  column_to_rownames("Pathway") %>%
  as.matrix()

storage.mode(mat_path) <- "numeric"

# sample x pathway
mat_rel <- t(mat_path)

# re-close after removing UNMAPPED
mat_rel <- mat_rel / rowSums(mat_rel)

# convert back to data frame
org_path_abun <- as.data.frame(mat_rel)
org_path_abun$Sample_raw <- rownames(org_path_abun)


### Rename pathway and select pathways
### Build pathway annotation table directly from the original pathway names
pathway_name_map <- path_abun_merged %>%
  distinct(Pathway) %>%
  transmute(
    Pathway_full = Pathway,
    ### Set shorter version of the pathway names
    # Pathway_short = ifelse(
    #   str_detect(Pathway, ":"),
    #   str_remove(Pathway, ":.*$"),
    #   Pathway
    # )
    Pathway_short = Pathway
  )

head(pathway_name_map)

### Match selected pathway IDs to full pathway names in abundance table
selected_pathway_full <- pathway_name_map %>%
  filter(Pathway_short %in% selected_pathway) %>%
  pull(Pathway_full)

selected_pathway_full
length(selected_pathway_full)

# Keep only selected pathways
select_path_abun <- org_path_abun %>%
  select(Sample_raw, all_of(selected_pathway_full))

### Remove excluded samples
samples_to_remove <- c(
  "NW016_2_2_all_merged_Abundance",
  "NW016_5_all_merged_Abundance",
  "NW046_5_all_merged_Abundance"
)

select_path_abun <- select_path_abun %>%
  filter(!Sample_raw %in% samples_to_remove)

### Convert table into long table
path_abun_long <- select_path_abun %>%
  pivot_longer(
    cols = -Sample_raw,
    names_to = "Pathway_full",
    values_to = "Abundance"
  )

# Update sample names in the long table
# Remove the ends of "_all_merged_Abundance" in the sample names
path_abun_long <- path_abun_long %>%
  mutate(
    Sample = Sample_raw %>%
      str_remove("_all_merged_Abundance$")
  ) %>%
  mutate(
    Sample = str_replace_all(Sample, "_", "-")
  )

### Add short pathway names for cleaner plot titles
path_abun_long <- path_abun_long %>%
  left_join(pathway_name_map, by = "Pathway_full")

head(path_abun_long)


### Mapping metadata (Age) to pathway abundance results
# Add Individual ID
path_abun_long <- path_abun_long %>%
  mutate(
    Individual = str_extract(Sample, "^[^-]+")
  )

# Individual-level Age metadata
metadata_age_table <- metadata_table %>%
  mutate(
    Individual = str_extract(Sample, "^[^-]+")
  ) %>%
  group_by(Individual) %>%
  summarise(
    Age = first(Age),
    .groups = "drop"
  )

# Join Age
path_abun_meta <- path_abun_long %>%
  left_join(metadata_age_table, by = "Individual")

head(path_abun_meta)



### Scatter plots for pathway abundance against Age
# pwy <- "UNINTEGRATED"
# pwy <- "DTDPRHAMSYN-PWY"

outdir <- "../../tmp_figures/MaAsLin3_Age_Pathway"
dir.create(outdir, recursive = TRUE, showWarnings = FALSE)

plot_list <- lapply(unique(path_abun_meta$Pathway_short), function(pwy) {
  
  df_sub <- path_abun_meta %>%
    filter(Pathway_short == pwy, !is.na(Age), !is.na(Abundance))
  
  if (nrow(df_sub) < 3) return(NULL)
  
  cor_res <- suppressWarnings(
    cor.test(df_sub$Age, df_sub$Abundance, method = "spearman", exact = FALSE)
  )
  
  pseudo <- 1e-6
  
  df_sub <- df_sub %>%
    mutate(Log10_Abundance = log10(Abundance + pseudo))
  
  cor_label <- paste0(
    "Spearman \u03C1 = ", round(cor_res$estimate, 3),
    "\nP = ", signif(cor_res$p.value, 3)
  )
  
  p <- ggplot(df_sub, aes(x = Age, y = Log10_Abundance)) +
    geom_point(size = 2.5, alpha = 0.8) +
    geom_smooth(method = "lm", se = FALSE, linewidth = 0.8) +
    annotate(
      "text",
      x = Inf, y = Inf,
      label = cor_label,
      hjust = 1.05, vjust = 1.2,
      size = 4
    ) +
    labs(
      title = str_wrap(pwy, width = 50),
      x = "Age (years)",
      y = expression(Log[10] * " relative pathway abundance")
    ) +
    theme_bw(base_size = 15) +
    theme(
      plot.title = element_text(face = "bold", hjust = 0.5, size = 15),
      axis.title = element_text(face = "bold"),
      axis.text = element_text(color = "black"),
      plot.margin = margin(8, 12, 8, 8)
    )

  ### Saving Figures
  outdir <- "../../tmp_figures/MaAsLin3_Age_Pathway"

  safe_name <- gsub("[^A-Za-z0-9_\\-]", "_", pwy)

  ggsave(
    filename = file.path(outdir, paste0(safe_name, "_Age_scatter.jpg")),
    plot = p,
    width = 6,
    height = 6,
    units = "in"
  )
  
  return(p)
  
})

plot_list <- plot_list[!sapply(plot_list, is.null)]
names(plot_list) <- unique(path_abun_meta$Pathway_short)[seq_along(plot_list)]

# Example
plot_list[[1]]


```



### Refined plots of associations between pathway abundance and age (Focused on interested pathways)
```{r}
library(dplyr)
library(tidyr)
library(readr)
library(stringr)
library(ggplot2)
library(ggrepel)
library(tibble)

# # SCFA-related Pathways
# selected_pathway <- c("CENTFERM-PWY", "PWY-5676", "PWY-5677", "P108-PWY", "PWY-8086", "P124-PWY", "FERMENTATION-PWY")

# # Vitamin synthesis Pathways
# selected_pathway <- c("1CMET2-PWY", "FOLSYN-PWY", "PWY-3841", "BIOTIN-BIOSYNTHESIS-PWY", "PWY-5005", "COBALSYN-PWY", "COA-PWY", "PANTO-PWY", "PANTOSYN-PWY", "NAD-BIOSYNTHESIS-II")

# Inflammation-related Pathways
selected_pathway <- c("LPSSYN-PWY", "KDO-NAGLIPASYN-PWY", "NAGLIPASYN-PWY", "LIPA-CORESYN-PWY", "OANTIGEN-PWY", "PEPTIDOGLYCANSYN-PWY", "PWY-6385", "PWY-6387", "PWY-6471", "AEROBACTINSYN-PWY", "COLANSYN-PWY", "LIPASYN-PWY")





### Read in sample metadata
metadata_file <- file.path("../../data/Metadata", "selected_northwood_map.txt")
metadata_table <- read_tsv(metadata_file, col_types = cols())

metadata_table <- metadata_table %>%
  rename(Sample = `#SampleID`) %>%
  select(Sample, Age, Sex, TimeNW, FIScore, Location)
head(metadata_table)


### Read in pathway abundance data
org_path_abun_file <- file.path(
  "../../data/Metagenomics/HUMAnN_func/Normalized_Tables",
  "merged_pathabundance_relab.tsv"
)

org_path_abun_table <- read.table(
  org_path_abun_file,
  header = TRUE,
  sep = "\t",
  comment.char = "",
  quote = "",
  check.names = FALSE
)

colnames(org_path_abun_table)[1] <- "Pathway"

### Remove stratified rows and unwanted rows
path_abun_merged <- org_path_abun_table %>%
  filter(!grepl("\\|", Pathway)) %>%
  filter(!Pathway %in% c("UNMAPPED"))

head(path_abun_merged)
nrow(path_abun_merged)


### Re-close abundance data after removing "UNMAPPED"
mat_path <- path_abun_merged %>%
  column_to_rownames("Pathway") %>%
  as.matrix()

storage.mode(mat_path) <- "numeric"

# sample x pathway
mat_rel <- t(mat_path)

# re-close after removing UNMAPPED
mat_rel <- mat_rel / rowSums(mat_rel)

# convert back to data frame
org_path_abun <- as.data.frame(mat_rel)
org_path_abun$Sample_raw <- rownames(org_path_abun)


### Rename pathway and select pathways
### Build pathway annotation table directly from the original pathway names
pathway_name_map <- path_abun_merged %>%
  distinct(Pathway) %>%
  transmute(
    Pathway_full = Pathway,
    Pathway_short = ifelse(
      str_detect(Pathway, ":"),
      str_remove(Pathway, ":.*$"),
      Pathway
    )
  )

head(pathway_name_map)

### Match selected pathway IDs to full pathway names in abundance table
selected_pathway_full <- pathway_name_map %>%
  filter(Pathway_short %in% selected_pathway) %>%
  pull(Pathway_full)

selected_pathway_full
length(selected_pathway_full)

# Keep only selected pathways
select_path_abun <- org_path_abun %>%
  select(Sample_raw, all_of(selected_pathway_full))

### Remove excluded samples
samples_to_remove <- c(
  "NW016_2_2_all_merged_Abundance",
  "NW016_5_all_merged_Abundance",
  "NW046_5_all_merged_Abundance"
)

select_path_abun <- select_path_abun %>%
  filter(!Sample_raw %in% samples_to_remove)

### Convert table into long table
path_abun_long <- select_path_abun %>%
  pivot_longer(
    cols = -Sample_raw,
    names_to = "Pathway_full",
    values_to = "Abundance"
  )

# Update sample names in the long table
# Remove the ends of "_all_merged_Abundance" in the sample names
path_abun_long <- path_abun_long %>%
  mutate(
    Sample = Sample_raw %>%
      str_remove("_all_merged_Abundance$")
  ) %>%
  mutate(
    Sample = str_replace_all(Sample, "_", "-")
  )

### Add short pathway names for cleaner plot titles
path_abun_long <- path_abun_long %>%
  left_join(pathway_name_map, by = "Pathway_full")

head(path_abun_long)


### Mapping metadata (Age) to pathway abundance results
# Add Individual ID
path_abun_long <- path_abun_long %>%
  mutate(
    Individual = str_extract(Sample, "^[^-]+")
  )

# Individual-level Age metadata
metadata_age_table <- metadata_table %>%
  mutate(
    Individual = str_extract(Sample, "^[^-]+")
  ) %>%
  group_by(Individual) %>%
  summarise(
    Age = first(Age),
    .groups = "drop"
  )

# Join Age
path_abun_meta <- path_abun_long %>%
  left_join(metadata_age_table, by = "Individual")

head(path_abun_meta)



### Scatter plots for pathway abundance against Age
# pwy <- "UNINTEGRATED"
# pwy <- "DTDPRHAMSYN-PWY"

outdir <- "../../tmp_figures/Interested_Age_Pathways"
dir.create(outdir, recursive = TRUE, showWarnings = FALSE)

plot_list <- lapply(unique(path_abun_meta$Pathway_short), function(pwy) {
  
  df_sub <- path_abun_meta %>%
    filter(Pathway_short == pwy, !is.na(Age), !is.na(Abundance))
  
  if (nrow(df_sub) < 3) return(NULL)
  
  cor_res <- suppressWarnings(
    cor.test(df_sub$Age, df_sub$Abundance, method = "spearman", exact = FALSE)
  )
  
  pseudo <- 1e-6
  
  df_sub <- df_sub %>%
    mutate(Log10_Abundance = log10(Abundance + pseudo))
  
  cor_label <- paste0(
    "Spearman \u03C1 = ", round(cor_res$estimate, 3),
    "\nP = ", signif(cor_res$p.value, 3)
  )
  
  p <- ggplot(df_sub, aes(x = Age, y = Log10_Abundance)) +
    geom_point(size = 2.5, alpha = 0.8) +
    geom_smooth(method = "lm", se = FALSE, linewidth = 0.8) +
    annotate(
      "text",
      x = Inf, y = Inf,
      label = cor_label,
      hjust = 1.05, vjust = 1.2,
      size = 4
    ) +
    labs(
      title = str_wrap(pwy, width = 45),
      x = "Age (years)",
      y = expression(Log[10] * " relative pathway abundance")
    ) +
    theme_bw(base_size = 15) +
    theme(
      plot.title = element_text(face = "bold", hjust = 0.5, size = 15),
      axis.title = element_text(face = "bold"),
      axis.text = element_text(color = "black"),
      plot.margin = margin(8, 12, 8, 8)
    )

  ### Saving Figures
  outdir <- "../../tmp_figures/Interested_Age_Pathways"

  safe_name <- gsub("[^A-Za-z0-9_\\-]", "_", pwy)

  ggsave(
    filename = file.path(outdir, paste0(safe_name, "_Age_scatter.jpg")),
    plot = p,
    width = 6,
    height = 6,
    units = "in"
  )
  
  return(p)
  
})

plot_list <- plot_list[!sapply(plot_list, is.null)]
names(plot_list) <- unique(path_abun_meta$Pathway_short)[seq_along(plot_list)]

# Example
plot_list[[1]]


```



### Study associations between drugs/medication results and pathway/taxa abundance
```{r}
library(dplyr)
library(maaslin3)

# ### Read in HUMANn pathway abundance table
# data_table_name <- file.path("../../data/Metagenomics/HUMAnN_func/Normalized_Tables", "merged_pathabundance_relab.tsv")
# 
# 
# org_data_table <- read.csv(data_table_name, sep = '\t', row.names = 1)
# nrow(org_data_table)
# clean_data_table <- org_data_table[!grepl("\\|", rownames(org_data_table)), ]
# nrow(clean_data_table)
# 
# # Remove "UNMAPPED" only
# clean_data_table <- clean_data_table[!rownames(clean_data_table) %in% c("UNMAPPED"), ]
# 
# # # Remove "UNINTEGRATED" only
# # clean_data_table <- clean_data_table[!rownames(clean_data_table) %in% c("UNINTEGRATED"), ]
# 
# # # Remove "UNMAPPED" and "UNINTEGRATED"
# # clean_data_table <- clean_data_table[!rownames(clean_data_table) %in% c("UNMAPPED", "UNINTEGRATED"), ]
# 
# # Update pathway names for better visualization
# rownames(clean_data_table) <- sub(":.*", "", rownames(clean_data_table))
# 
# nrow(clean_data_table)
# data_table <- t(clean_data_table)
# ncol(data_table)
# 
# 
# ### Filtering on Pathway Prevalence
# # Calculate pathway prevalence across samples
# prev <- colMeans(data_table > 0, na.rm = TRUE)
# 
# # Keep pathways with prevalence > 0.1
# prev_cut <- 0.10
# keep_pathways <- prev > prev_cut
# 
# # Filter pathways
# data_table <- data_table[, keep_pathways, drop = FALSE]
# 
# # Check number of retained pathways
# ncol(data_table)
# 
# 
# # Remove Duplicate Samples and Sample without metadata
# data_table <- data_table[!rownames(data_table) %in% c("NW016_2_2_all_merged_Abundance", "NW016_5_all_merged_Abundance", "NW046_5_all_merged_Abundance"), ]
# data_table[1:5, 1:5]
# nrow(data_table)





### Read in MetaPhlAn taxa abundance table
data_table_name <- file.path("../../data/Metagenomics/MetaPhlAn_taxa", "merged_abundance_table.txt")
org_data_table <- read.csv(data_table_name, sep = '\t', row.names = 1, comment.char = "#")
nrow(org_data_table)

# # Select rows with species level annotation
# clean_data_table <- org_data_table[
#   grepl("\\|s__", rownames(org_data_table)) &
#   !grepl("\\|t__", rownames(org_data_table)),
# ]

# Select rows with genus level annotation
clean_data_table <- org_data_table[
  grepl("\\|g__", rownames(org_data_table)) &
  !grepl("\\|s__", rownames(org_data_table)),
]

# # Select rows with family level annotation
# clean_data_table <- org_data_table[
#   grepl("\\|f__", rownames(org_data_table)) &
#   !grepl("\\|g__", rownames(org_data_table)),
# ]

# # Select rows with phylum level annotation
# clean_data_table <- org_data_table[
#   grepl("\\|p__", rownames(org_data_table)) &
#   !grepl("\\|c__", rownames(org_data_table)),
# ]

nrow(clean_data_table)
data_table <- t(clean_data_table)

# Remove Duplicate Samples and Sample without metadata
data_table <- data_table[!rownames(data_table) %in% c("NW016_2_2_all", "NW016_5_all", "NW046_5_all"), ]
# colnames(data_table) <- sub(".*\\|s__", "", colnames(data_table))
colnames(data_table) <- sub(".*\\|g__", "", colnames(data_table))
# colnames(data_table) <- sub(".*\\|f__", "", colnames(data_table))
# colnames(data_table) <- sub(".*\\|p__", "", colnames(data_table))


data_table[1:5, 1:5]
nrow(data_table)
ncol(data_table)


### Read in metadata table
metadata_name <- file.path("../../data/Metadata", "patients_master_metadata.txt")
org_metadata <- read.csv(metadata_name, sep = '\t', row.names = 1)
# Remove sample without metadata
metadata <- org_metadata[!rownames(org_metadata) %in% c("NW046"), ]
head(metadata)
nrow(metadata)

# Matching ID names
# 1. Extract Individual ID
data_id <- sub("_.*", "", rownames(data_table))
meta_id <- rownames(metadata)
# 2. Matching
match_idx <- match(meta_id, data_id)
match_idx
# 3. Remove unmatched ID
keep <- !is.na(match_idx)
metadata_matched <- metadata[keep, ]
match_idx <- match_idx[keep]
# 4. Replace/Update rownames of the metadata
rownames(metadata_matched) <- rownames(data_table)[match_idx]
metadata <- metadata_matched
head(metadata)
head(data_table)



### Factor the categorical variables in metadata
metadata$Sex <- 
  factor(metadata$Sex, levels = c('Male', 'Female'))

metadata$Location <- 
  factor(metadata$Location, levels = c('Manor', 'Center'))



### Factor the categorical variables in medication columns in the metadata
binary_vars <- c(
  "anemia", "arthritis", "atherosclerosis", "copd", "dementia",
  "depression", "diabetes", "dyslipidemia", "hypertension",
  "hypothyroidism", "osteoporosis", "renal_disease",
  "ace_inhibitors", "acetaminophen", "aaps", "benzodiazepines",
  "beta_blockers", "antiplatelets", "ccbs", "diuretics", "insulin",
  "iron_supplements", "laxatives", "levothyroxine", "metformin",
  "nitrates", "opioids", "ppis", "ssris", "statins", "tcas",
  "vitb12", "vitd"
)

metadata <- metadata %>%
  mutate(
    across(
      all_of(binary_vars),
      ~ factor(.x, levels = c(0, 1))
    )
  )

# Check dataframes
data_table[1:5, 1:5]
metadata[1:5, ]


### Running MaAsLin3 against diseases
# formula = '~ Sex + Age + TimeNW + FIScore + Location',

# set.seed(1)
# fit_out <- maaslin3(input_data = data_table,
#                     input_metadata = metadata,
#                     output = 'MaAsLin3_pathway_rm_unmap_disease_with_meta_output',
#                     formula = '~ Sex + Age + TimeNW + FIScore + Location + anemia + arthritis + atherosclerosis + copd + dementia + depression + diabetes + dyslipidemia + hypertension + hypothyroidism + osteoporosis + renal_disease',
#                     normalization = 'TSS',
#                     transform = 'LOG',
#                     augment = TRUE,
#                     standardize = TRUE,
#                     max_significance = 0.1,
#                     median_comparison_abundance = TRUE,
#                     median_comparison_prevalence = FALSE,
#                     max_pngs = 250,
#                     cores = 4)


set.seed(1)
fit_out <- maaslin3(input_data = data_table,
                    input_metadata = metadata,
                    #output = 'MaAsLin3_pathway_rm_unmap_disease_with_age_output',
                    output = 'MaAsLin3_MG_genus_disease_without_metadata_output',
                    formula = '~ anemia + arthritis + atherosclerosis + copd + dementia + depression + diabetes + dyslipidemia + hypertension + hypothyroidism + osteoporosis + renal_disease',
                    normalization = 'TSS',
                    transform = 'LOG',
                    augment = TRUE,
                    standardize = TRUE,
                    max_significance = 0.1,
                    median_comparison_abundance = TRUE,
                    median_comparison_prevalence = FALSE,
                    max_pngs = 250,
                    cores = 4)


### Running MaAsLin3 against medicine
set.seed(1)
fit_out <- maaslin3(input_data = data_table,
                    input_metadata = metadata,
                    # output = 'MaAsLin3_pathway_rm_unmap_medicine_with_age_output',
                    output = 'MaAsLin3_MG_genus_medicine_without_metadata_output',
                    formula = '~ ace_inhibitors + acetaminophen + aaps + benzodiazepines + beta_blockers + antiplatelets + ccbs + diuretics + insulin + iron_supplements + laxatives + levothyroxine + metformin + nitrates + opioids + ppis + ssris + statins + tcas + vitb12 + vitd',
                    normalization = 'TSS',
                    transform = 'LOG',
                    augment = TRUE,
                    standardize = TRUE,
                    max_significance = 0.1,
                    median_comparison_abundance = TRUE,
                    median_comparison_prevalence = FALSE,
                    max_pngs = 250,
                    cores = 4)


### Running MaAsLin3 against metadata
set.seed(1)
fit_out <- maaslin3(input_data = data_table,
                    input_metadata = metadata,
                    #output = 'MaAsLin3_pathway_rm_unmap_with_meta_output',
                    output = 'MaAsLin3_MG_genus_metadata_output',
                    formula = '~ Sex + Age + TimeNW + FIScore + Location',
                    normalization = 'TSS',
                    transform = 'LOG',
                    augment = TRUE,
                    standardize = TRUE,
                    max_significance = 0.1,
                    median_comparison_abundance = TRUE,
                    median_comparison_prevalence = FALSE,
                    max_pngs = 250,
                    cores = 4)

```






Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
