---
title: "MaAsLin3_16S_Analysis"
author: "Zhongzhi"
date: "2026-04-09"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```



### 16S Taxa Units MaSaLin3 Analysis (Metadata)
```{r pressure, echo=FALSE}
setwd("C:/Users/dlszz/OneDrive - Dalhousie University/Nursing_House_Project/Git_Project/northwood-microbiome/code/MG_analysis")

library(dplyr)
library(stringr)
library(tidyr)
library(tibble)
library(maaslin3)

### Read in 16S ASV feature table
#data_table_name <- file.path("../../tmp_figures/16S_MaAsLin3_data", "table_bleed_decon.tsv")
# Read in updated 16S ASV feature table
data_table_name <- file.path("../../tmp_figures/16S_MaAsLin3_data/update_16S_0421", "feature-table.tsv")

org_data_table <- read.csv(data_table_name, sep = '\t', row.names = 1, skip = 1, check.names = FALSE)
nrow(org_data_table)
org_data_table[1:5, 1:5]
nrow(org_data_table)
ncol(org_data_table)

# # Remove Duplicate Samples and Sample without metadata
# data_table <- org_data_table[, !colnames(org_data_table) %in% c("NW046-1", "NW046-5")]
data_table <- org_data_table
data_table[1:5, 1:5]
nrow(data_table)
ncol(data_table)



### Read in ASV taxonomic mappings
taxa_map_name <- file.path("../../tmp_figures/16S_MaAsLin3_data/update_16S_0421", "taxonomy.tsv")
taxa_map_table <- read.csv(taxa_map_name, sep = '\t', header = TRUE, check.names = FALSE)
head(taxa_map_table)
# New columns for taxonomic information
taxa_map_table <- taxa_map_table %>%
  mutate(.row_id = row_number()) %>%
  # Seperate the Taxon column
  separate_rows(Taxon, sep = ";") %>%
  mutate(Taxon = str_trim(Taxon)) %>%
  mutate(
    rank = str_extract(Taxon, "^[dpcfogs]__"),
    value = str_replace(Taxon, "^[dpcfogs]__", ""),
    value = if_else(is.na(value) | value == "", "Unknown", value)
  ) %>%
  # Taxonomy ranks
  filter(rank %in% c("d__", "p__", "c__", "o__", "f__", "g__", "s__")) %>%
  # Wide table
  pivot_wider(
    id_cols = c(.row_id, `Feature ID`, Confidence),
    names_from = rank,
    values_from = value,
    values_fn = dplyr::first
  ) %>%
  # Rename Columns
  rename(
    Domain  = `d__`,
    Phylum  = `p__`,
    Class   = `c__`,
    Order   = `o__`,
    Family  = `f__`,
    Genus   = `g__`,
    Species = `s__`
  ) %>%
  mutate(across(c(Domain, Phylum, Class, Order, Family, Genus, Species),
                ~ if_else(is.na(.x) | .x == "", "Unknown", .x))) %>%
  select(-.row_id)

taxa_map_table[1:5, ]
colnames(taxa_map_table)[1] <- "ASV"
taxa_map_table[1:5, ]



# ### Read in 16S metadata
# metadata_name <- file.path("../../tmp_figures/16S_MaAsLin3_data", "northwood_map.txt")
### Read in update 16S metadata
metadata_name <- file.path("../../tmp_figures/16S_MaAsLin3_data/update_16S_0421", "update_northwood_map.txt")
org_metadata <- read.csv(metadata_name, sep = '\t', header = TRUE)
names(org_metadata)[1] <- "SampleID"
head(org_metadata)

# Remove sample without metadata
measured_samples <- colnames(data_table)
metadata <- org_metadata %>% filter(SampleID %in% measured_samples)
head(metadata)
nrow(metadata)


### Factor the categorical variables in metadata
metadata$Sex <- 
  factor(metadata$Sex, levels = c('Male', 'Female'))

metadata$Location <- 
  factor(metadata$Location, levels = c('Manor', 'Center'))

metadata$FICat <- 
  factor(metadata$FICat, levels = c('3', '4'))

# Check dataframes
data_table[1:5, 1:5]
metadata[1:5, ]





###### Perform MaAsLin3_16S_Analysis
### Add specific level taxonomic annotation to the feature table
# level <- "Phylum"
level <- "Family"
# level <- "Genus"
# level <- "Species"
# level <- "ASV_ID"
# taxa_map_table$ASV_ID <- taxa_map_table$ASV


data_table <- data_table %>%
  rownames_to_column(var = "ASV")
nrow(taxa_map_table)
select_taxa_map_table <- taxa_map_table %>% select("ASV", all_of(level)) %>% distinct()
nrow(select_taxa_map_table)
table(select_taxa_map_table[[level]])


anno_data_table <- data_table %>% 
  left_join(select_taxa_map_table, by = "ASV") %>%
  mutate(!!level := if_else(is.na(.data[[level]]) | .data[[level]] == "", "Unknown", .data[[level]]))

### Summarise OTU abundance at specific level
sample_cols <- setdiff(colnames(anno_data_table), c("ASV", level))
sum_anno_data_table <- anno_data_table %>%
  select(all_of(level), all_of(sample_cols)) %>%
  group_by(.data[[level]]) %>%
  summarise(across(all_of(sample_cols), ~ sum(as.numeric(.x), na.rm = TRUE)), .groups = "drop") %>%
  arrange(desc(rowSums(across(all_of(sample_cols)))))

sum_anno_data_table

# ### Output family-level taxonomic profilings
# family_level_taxa_file <- file.path("../../tmp_figures/16S_MaAsLin3_data", "16S_Family_Profiling.tsv")
# family_level_taxa_file <- file.path("../../tmp_figures/16S_MaAsLin3_data", "all_samples_16S_Family_Profiling.tsv")
# family_level_taxa_file <- file.path("../../tmp_figures/16S_MaAsLin3_data", "all_samples_16S_Genus_Profiling.tsv")
# write.table(sum_anno_data_table, family_level_taxa_file, sep = "\t", quote = FALSE, row.names = FALSE, col.names = TRUE)


### Convert taxa table format for MaAsLin input
sum_df <- as.data.frame(sum_anno_data_table, check.names = FALSE)
rownames(sum_df) <- sum_df[[level]]
sum_df[[level]] <- NULL
final_data_table <- as.data.frame(t(sum_df), check.names = FALSE)
final_data_table[1:5, ]
nrow(final_data_table)

metadata[1:5, ]
nrow(metadata)

final_metadata <- metadata %>%
  mutate(
    Individual = str_extract(SampleID, "^[^-]+")  # 提取到第一个 '-' 之前
  ) %>%
  column_to_rownames("SampleID")
final_metadata[1:5, ]
final_data_table[1:5, ]
ncol(final_data_table)


### Running MaAsLin3
set.seed(1)
fit_out <- maaslin3(input_data = final_data_table,
                    input_metadata = final_metadata,
                    #output = 'MaAsLin3_16S_genus_test_output',
                    output = 'MaAsLin3_16S_family_update_output',
                    formula = '~ Sex + Age +
                        TimeNW + FIScore + Location + (1|Individual)',
                    normalization = 'TSS',
                    transform = 'LOG',
                    coef_plot_vars = c("Age", "Location Center"),
                    heatmap_vars = c("FIScore", "TimeNW", "Sex Female"),
                    augment = TRUE,
                    standardize = TRUE,
                    max_significance = 0.1,
                    median_comparison_abundance = TRUE,
                    median_comparison_prevalence = FALSE,
                    max_pngs = 250,
                    cores = 4)


```



### 16S Taxa Units (Disease and medication)
```{r}
library(dplyr)
library(tibble)
library(maaslin3)
library(stringr)

### Read in 16S taxonomic results
#level <- "Family"
#family_level_taxa_file <- file.path("../../tmp_figures/16S_MaAsLin3_data", "all_samples_16S_Family_Profiling.tsv")
level <- "Genus"
family_level_taxa_file <- file.path("../../tmp_figures/16S_MaAsLin3_data", "all_samples_16S_Genus_Profiling.tsv")

sum_anno_data_table <- read.table(family_level_taxa_file, sep = "\t", header = TRUE, check.names = FALSE)

### Convert taxa table format for MaAsLin input
sum_df <- as.data.frame(sum_anno_data_table, check.names = FALSE)
rownames(sum_df) <- sum_df[[level]]
sum_df[[level]] <- NULL
final_data_table <- as.data.frame(t(sum_df), check.names = FALSE)
final_data_table[1:5, ]
nrow(final_data_table)
# Remove sample without metadata
final_data_table <- final_data_table[
  !grepl("NW046", rownames(final_data_table)),
]
head(final_data_table)
nrow(final_data_table)


### Read in sample metadata
metadata_name <- file.path("../../tmp_figures/16S_MaAsLin3_data/update_16S_0421", "update_northwood_map.txt")
org_metadata <- read.csv(metadata_name, sep = '\t', header = TRUE)
names(org_metadata)[1] <- "SampleID"
head(org_metadata)
nrow(org_metadata)

# Remove sample without metadata
measured_samples <- rownames(final_data_table)
metadata <- org_metadata %>% filter(SampleID %in% measured_samples)
head(metadata)
nrow(metadata)

final_metadata <- metadata %>%
  mutate(
    Individual = str_extract(SampleID, "^[^-]+")  # 提取到第一个 '-' 之前
  ) %>%
  column_to_rownames("SampleID")
final_metadata[1:5, ]
final_data_table[1:5, ]
ncol(final_data_table)

metadata[1:5, ]
nrow(metadata)

final_metadata <- metadata %>%
  mutate(
    Individual = str_extract(SampleID, "^[^-]+")  # 提取到第一个 '-' 之前
  ) %>%
  column_to_rownames("SampleID")
final_metadata[1:5, ]

# Remove sample without metadata
final_metadata <- final_metadata[
  !grepl("NW046", rownames(final_metadata)),
]
nrow(final_metadata)



### Read in individual medication metadata
med_metadata_name <- file.path("../../data/Metadata", "patients_master_metadata.txt")
org_med_metadata <- read.csv(med_metadata_name, sep = '\t', row.names = 1)
# Remove sample without metadata
med_metadata <- org_med_metadata [!rownames(org_med_metadata) %in% c("NW046"), ]
head(med_metadata)
nrow(med_metadata)
selected_med_metadata <- med_metadata %>%
  rownames_to_column(var = "Individual") %>%
  select(Individual, anemia:last_col())
head(selected_med_metadata)

# Matching medication metadata to basic metadata by individual name
head(final_metadata)
final_metadata <- final_metadata %>%
  rownames_to_column(var = "Sample") %>%
  left_join(selected_med_metadata, by = "Individual") %>%
  column_to_rownames(var = "Sample")
head(final_metadata)


### Factor the categorical variables in metadata
final_metadata$Sex <- 
  factor(metadata$Sex, levels = c('Male', 'Female'))

final_metadata$Location <- 
  factor(metadata$Location, levels = c('Manor', 'Center'))



### Factor the categorical variables in medication columns in the metadata
binary_vars <- c(
  "anemia", "arthritis", "atherosclerosis", "copd", "dementia",
  "depression", "diabetes", "dyslipidemia", "hypertension",
  "hypothyroidism", "osteoporosis", "renal_disease",
  "ace_inhibitors", "acetaminophen", "aaps", "benzodiazepines",
  "beta_blockers", "antiplatelets", "ccbs", "diuretics", "insulin",
  "iron_supplements", "laxatives", "levothyroxine", "metformin",
  "nitrates", "opioids", "ppis", "ssris", "statins", "tcas",
  "vitb12", "vitd"
)

final_metadata <- final_metadata %>%
  mutate(
    across(
      all_of(binary_vars),
      ~ factor(.x, levels = c(0, 1))
    )
  )

# Check dataframes
data_table[1:5, 1:5]
final_metadata[1:5, ]



### Running MaAsLin3
set.seed(1)
fit_out <- maaslin3(input_data = final_data_table,
                    input_metadata = final_metadata,
                    output = 'MaAsLin3_16S_genus_metadata_output',
                    formula = '~ Sex + Age +
                        TimeNW + FIScore + Location + (1|Individual)',
                    normalization = 'TSS',
                    transform = 'LOG',
                    augment = TRUE,
                    standardize = TRUE,
                    max_significance = 0.1,
                    median_comparison_abundance = TRUE,
                    median_comparison_prevalence = FALSE,
                    max_pngs = 250,
                    cores = 4)

set.seed(1)
fit_out <- maaslin3(input_data = final_data_table,
                    input_metadata = final_metadata,
                    output = 'MaAsLin3_16S_genus_disease_without_metadata_output',
                    formula = '~ anemia + arthritis + atherosclerosis + copd + dementia + depression + diabetes + dyslipidemia + hypertension + hypothyroidism + osteoporosis + renal_disease + (1|Individual)',
                    normalization = 'TSS',
                    transform = 'LOG',
                    augment = TRUE,
                    standardize = TRUE,
                    max_significance = 0.1,
                    median_comparison_abundance = TRUE,
                    median_comparison_prevalence = FALSE,
                    max_pngs = 250,
                    cores = 4)


### Running MaAsLin3
set.seed(1)
fit_out <- maaslin3(input_data = final_data_table,
                    input_metadata = final_metadata,
                    output = 'MaAsLin3_16S_genus_medication_without_metadata_output',
                    formula = '~ ace_inhibitors + acetaminophen + aaps + benzodiazepines + beta_blockers + antiplatelets + ccbs + diuretics + insulin + iron_supplements + laxatives + levothyroxine + metformin + nitrates + opioids + ppis + ssris + statins + tcas + vitb12 + vitd + (1|Individual)',
                    normalization = 'TSS',
                    transform = 'LOG',
                    augment = TRUE,
                    standardize = TRUE,
                    max_significance = 0.1,
                    median_comparison_abundance = TRUE,
                    median_comparison_prevalence = FALSE,
                    max_pngs = 250,
                    cores = 4)

```



### 16S- and MG-based taxonomic Profiling (Family Level. No figure generated, but serve as input for following sections)
```{r}
library(dplyr)
library(tidyr)
library(stringr)
library(ggplot2)
library(RColorBrewer)

# setwd()
###### Read in Metagenomics-based taxonomic profilings
### MG taxonomic profiling from MetaPhlAn4
MG_tax_file <- file.path("../../data/Metagenomics/MetaPhlAn_taxa", "merged_abundance_table.txt")
df <- read.delim(MG_tax_file, comment.char = "#", check.names = FALSE)

# Column Name Checking
stopifnot("clade_name" %in% colnames(df))
head(df)
nrow(df)

# Keep Family-level Taxa Anno
df_fam <- df %>%
  filter(str_detect(clade_name, "\\|f__")) %>%
  filter(!str_detect(clade_name, "\\|g__"))
nrow(df_fam)

# Extract Family Name
df_fam <- df_fam %>%
  mutate(Family = str_replace(clade_name, ".*\\|f__", "")) %>%
  select(Family, everything(), -clade_name)

head(df_fam)
tail(df_fam)
ncol(df_fam)



###### Read in 16S-based taxonomic profilings
S16_family_taxa_file <- file.path("../../data/Metagenomics/16S_related", "all_samples_16S_Family_Profiling.tsv")
S16_family_taxa_profile <- read.table(S16_family_taxa_file, header = TRUE, sep = "\t", check.names = FALSE)
head(S16_family_taxa_profile)
S16_samples <- colnames(S16_family_taxa_profile)

### Extract 16S samples with MG data
# Get MG sample names
MG_df_fam <- df_fam
MG_samples <- colnames(MG_df_fam)
MG_samples
MG_samples <- MG_samples[MG_samples != "Family"]
# Convert MG sample names to 16S-style sample IDs
MG_samples_16S_style <- MG_samples %>%
  gsub("_all$", "", .) %>%   # remove trailing "_all"
  gsub("_", "-", .)          # replace "_" with "-"

MG_samples_16S_style

### Select 16S samples that having MG-based profilings
selected_16S_cols <- c("Family", intersect(S16_samples, MG_samples_16S_style))
selected_16S_family_taxa_profile <- S16_family_taxa_profile[, selected_16S_cols]
# Check results
colnames(selected_16S_family_taxa_profile)
dim(selected_16S_family_taxa_profile)


top_n = 15
### Top 15 Families and Other Families in MG samples
# Convert into long table
MG_long_fam <- MG_df_fam %>%
  pivot_longer(
    cols = -Family,
    names_to = "Sample",
    values_to = "Abundance"
  ) %>%
  mutate(
    Abundance = as.numeric(Abundance),
    Abundance = replace_na(Abundance, 0)
  )

# Top families in MG data
MG_top_fam <- MG_long_fam %>%
  group_by(Family) %>%
  summarise(total = sum(Abundance, na.rm = TRUE), .groups = "drop") %>%
  arrange(desc(total)) %>%
  slice_head(n = top_n) %>%
  pull(Family)

# Combine left families into Other taxa
MG_plot_df <- MG_long_fam  %>%
  mutate(Family_plot = if_else(Family %in% MG_top_fam, Family, "Other")) %>%
  group_by(Sample, Family_plot) %>%
  summarise(Abundance = sum(Abundance, na.rm = TRUE), .groups = "drop")

head(MG_plot_df)
# Convert MG sample names to 16S-style sample IDs
MG_plot_df <- MG_plot_df %>%
  mutate(
    Sample = Sample %>%
      gsub("_all$", "", .) %>%
      gsub("_", "-", .)
  )

head(MG_plot_df)
tail(MG_plot_df)




### Top 15 Families and Other Families in 16S samples
head(selected_16S_family_taxa_profile)

# Convert into long table
S16_long_fam <- selected_16S_family_taxa_profile %>%
  pivot_longer(
    cols = -Family,
    names_to = "Sample",
    values_to = "Abundance"
  ) %>%
  mutate(
    Abundance = as.numeric(Abundance),
    Abundance = replace_na(Abundance, 0)
  )

# Calculate family relative abundance in the samples
S16_long_fam <- S16_long_fam %>%
  group_by(Sample) %>%
  mutate(Rel_Abundance = Abundance / sum(Abundance, na.rm = TRUE)) %>%
  ungroup()

# Top families in 16S data
S16_top_fam <- S16_long_fam %>%
  group_by(Family) %>%
  summarise(total = sum(Rel_Abundance, na.rm = TRUE), .groups = "drop") %>%
  arrange(desc(total)) %>%
  slice_head(n = top_n) %>%
  pull(Family)
S16_top_fam

# Combine left families into Other taxa
S16_plot_df <- S16_long_fam  %>%
  mutate(Family_plot = if_else(Family %in% S16_top_fam, Family, "Other")) %>%
  group_by(Sample, Family_plot) %>%
  summarise(Abundance = sum(Abundance, na.rm = TRUE), Rel_Abundance = sum(Rel_Abundance, na.rm = TRUE), .groups = "drop")

head(S16_plot_df)


### Get the name of Top Families in both MG and 16S
top_fam_union <- union(MG_top_fam, S16_top_fam)


###### Regenerate 16S and MG tables
### MG tables
# Convert into long table
update_MG_long_fam <- MG_df_fam %>%
  pivot_longer(
    cols = -Family,
    names_to = "Sample",
    values_to = "Abundance"
  ) %>%
  mutate(
    Abundance = as.numeric(Abundance),
    Abundance = replace_na(Abundance, 0)
  )
# Combine left families into Other taxa
update_MG_plot_df <- update_MG_long_fam  %>%
  mutate(Family_plot = if_else(Family %in% top_fam_union, Family, "Other")) %>%
  group_by(Sample, Family_plot) %>%
  summarise(Abundance = sum(Abundance, na.rm = TRUE), .groups = "drop")

### S16 tables
# Convert into long table
update_S16_long_fam <- selected_16S_family_taxa_profile %>%
  pivot_longer(
    cols = -Family,
    names_to = "Sample",
    values_to = "Abundance"
  ) %>%
  mutate(
    Abundance = as.numeric(Abundance),
    Abundance = replace_na(Abundance, 0)
  )

update_S16_long_fam <- update_S16_long_fam %>%
  group_by(Sample) %>%
  mutate(Rel_Abundance = Abundance / sum(Abundance, na.rm = TRUE)) %>%
  ungroup()

# Combine left families into Other taxa
update_S16_plot_df <- update_S16_long_fam  %>%
  mutate(Family_plot = if_else(Family %in% top_fam_union, Family, "Other")) %>%
  group_by(Sample, Family_plot) %>%
  summarise(
    Abundance = sum(Abundance, na.rm = TRUE),
    Rel_Abundance = sum(Rel_Abundance, na.rm = TRUE),
    .groups = "drop"
  )

head(update_S16_plot_df)
### Calculate relative abundance


###### Barplots
### Using a same color set for both 16S data and MG data
all_taxa <- unique(c(sort(unique(top_fam_union)), "Other"))
# pal <- colorRampPalette(brewer.pal(12, "Set3"))(length(all_taxa))
# names(pal) <- all_taxa
# # Reorder palette to match all_taxa order
# pal <- pal[all_taxa]
# names(pal) <- all_taxa
# pal["Other"] <- "grey70"   # force grey (safe even if already grey)

taxa_no_other <- setdiff(all_taxa, "Other")

base_cols <- c(
  brewer.pal(8, "Set2"),
  brewer.pal(12, "Set3"),
  brewer.pal(8, "Dark2"),
  brewer.pal(9, "Set1")
)

bad_cols <- c(
  "#D9D9D9", # grey, too close to Other
  "#B3B3B3", # grey, too close to Other
  "#FFFFB3", # very pale yellow
  "#B3DE69", # green, too close to other greens
  "#FFED6F"  # yellow, sometimes too close to other yellows
)

base_cols <- base_cols[!toupper(base_cols) %in% toupper(bad_cols)]

pal <- base_cols[seq_along(taxa_no_other)]
names(pal) <- taxa_no_other

pal <- c(pal, Other = "grey70")
pal <- pal[all_taxa]


# Fix Taxon order (optional but recommended)
update_MG_plot_df <- update_MG_plot_df %>%
  mutate(Family_plot = factor(Family_plot, levels = all_taxa))
update_S16_plot_df <- update_S16_plot_df %>%
  mutate(Family_plot = factor(Family_plot, levels = all_taxa))


### Plot (MG)
update_MG_plot_df <- update_MG_plot_df %>% mutate(Sample = Sample %>%
      gsub("_all$", "", .) %>%
      gsub("_", "-", .))

p_fam_MG <- ggplot(update_MG_plot_df, aes(x = Sample, y = Abundance, fill = Family_plot)) +
  geom_col(width = 0.9) +
  scale_fill_manual(values = pal) +
  labs(x = "Sample", y = "Relative abundance (%)", fill = "Family") +
  theme_bw(base_size = 12) +
  theme(
    axis.text.x = element_text(angle = 75, hjust = 1, vjust = 1),
    panel.grid.major.x = element_blank()
  )
print(p_fam_MG)

### Saving MG Family Figures
# figure_file <- file.path("../../tmp_figures/Taxonomics_Plot/", "recolored_MG_based_family_level_tax_profile.jpg")
# ggsave(figure_file, p_fam_MG, width = 12, height = 6, dpi = 300)
# figure_file <- file.path("../../tmp_figures/Taxonomics_Plot/", "recolored_MG_based_family_level_tax_profile.pdf")
# ggsave(figure_file, p_fam_MG, width = 12, height = 6)
# figure_file <- file.path("../../tmp_figures/Taxonomics_Plot", "recolored_MG_based_family_level_tax_profile.svg")
# ggsave(figure_file, p_fam_MG, width = 12, height = 6)


### Plot (16S)
p_fam_S16 <- ggplot(update_S16_plot_df, aes(x = Sample, y = Rel_Abundance*100, fill = Family_plot)) +
  geom_col(width = 0.9) +
  scale_fill_manual(values = pal) +
  labs(x = "Sample", y = "Relative abundance (%)", fill = "Family") +
  theme_bw(base_size = 12) +
  theme(
    axis.text.x = element_text(angle = 75, hjust = 1, vjust = 1),
    panel.grid.major.x = element_blank()
  )
print(p_fam_S16)

### Saving S16 Family Figures
# figure_file <- file.path("../../tmp_figures/Taxonomics_Plot/", "recolored_S16_based_family_level_tax_profile.jpg")
# ggsave(figure_file, p_fam_S16, width = 12, height = 6, dpi = 300)
# figure_file <- file.path("../../tmp_figures/Taxonomics_Plot/", "recolored_S16_based_family_level_tax_profile.pdf")
# ggsave(figure_file, p_fam_S16, width = 12, height = 6)
# figure_file <- file.path("../../tmp_figures/Taxonomics_Plot", "recolored_S16_based_family_level_tax_profile.svg")
# ggsave(figure_file, p_fam_S16, width = 12, height = 6)


# Saving family level plot dataframe
update_MG_fam_plot_df <- update_MG_plot_df
update_S16_fam_plot_df <- update_S16_plot_df
fam_pal <- pal

# ###### 16S and MG Grouped Plots
# library(dplyr)
# library(ggplot2)
# 
# ### Make sure MG sample names match 16S style
# update_MG_plot_df <- update_MG_plot_df %>%
#   mutate(
#     Sample = Sample %>%
#       gsub("_all$", "", .) %>%
#       gsub("_", "-", .)
#   )
# 
# ### If MG already has relative abundance in percent, rename it clearly;
# ### otherwise replace MG_abundance_pct with the correct relative abundance column
# MG_plot_ready <- update_MG_plot_df %>%
#   mutate(
#     Method = "Metagenomics",
#     Abundance_pct = Abundance   # 推荐这样；如果你现在只有百分比原值，再相应替换
#   ) %>%
#   select(Sample, Family_plot, Method, Abundance_pct)
# 
# S16_plot_ready <- update_S16_plot_df %>%
#   mutate(
#     Method = "16S rRNA",
#     Abundance_pct = Rel_Abundance * 100
#   ) %>%
#   select(Sample, Family_plot, Method, Abundance_pct)
# 
# ### Combine two tables
# plot_df <- bind_rows(MG_plot_ready, S16_plot_ready)
# 
# ### Fix sample order
# sample_order <- sort(unique(plot_df$Sample))
# plot_df <- plot_df %>%
#   mutate(
#     Sample = factor(Sample, levels = sample_order),
#     Method = factor(Method, levels = c("16S rRNA", "Metagenomics")),
#     Family_plot = factor(Family_plot, levels = all_taxa)
#   )
# 
# ### Combined plot
# p_combined <- ggplot(plot_df, aes(x = Sample, y = Abundance_pct, fill = Family_plot)) +
#   geom_col(width = 0.9) +
#   facet_wrap(~ Method, ncol = 1) +
#   scale_fill_manual(values = pal) +
#   labs(x = "Sample", y = "Relative abundance (%)", fill = "Family") +
#   theme_bw(base_size = 12) +
#   theme(
#     axis.text.x = element_text(angle = 75, hjust = 1, vjust = 1),
#     panel.grid.major.x = element_blank(),
#     strip.text = element_text(face = "bold"),
#     legend.position = "right"
#   )
# 
# print(p_combined)
# 
# # figure_file <- file.path("../../tmp_figures/", "recolored_S16_MG_family_level_tax_profile.jpg")
# # ggsave(figure_file, p_combined, width = 12, height = 8, dpi = 300)
```



### 16S- and MG-based taxonomic Profiling (No figure generated, but serve as input for following sections)
```{r}
library(dplyr)
library(tidyr)
library(stringr)
library(ggplot2)
library(RColorBrewer)

#setwd()

###### Read in Metagenomics-based taxonomic profilings
### MG taxonomic profiling from MetaPhlAn4
MG_tax_file <- file.path("../../data/Metagenomics/MetaPhlAn_taxa", "merged_abundance_table.txt")
df <- read.delim(MG_tax_file, comment.char = "#", check.names = FALSE)

# Column Name Checking
stopifnot("clade_name" %in% colnames(df))
head(df)
nrow(df)

# Keep Genus-level Taxa Anno
df_fam <- df %>%
  filter(str_detect(clade_name, "\\|g__")) %>%
  filter(!str_detect(clade_name, "\\|s__"))
nrow(df_fam)

# Extract Genus Name
df_fam <- df_fam %>%
  mutate(Family = str_replace(clade_name, ".*\\|g__", "")) %>%
  select(Family, everything(), -clade_name)

head(df_fam)
tail(df_fam)
ncol(df_fam)



###### Read in 16S-based taxonomic profilings
S16_family_taxa_file <- file.path("../../data/Metagenomics/16S_related", "all_samples_16S_Genus_Profiling.tsv")
S16_family_taxa_profile <- read.table(S16_family_taxa_file, header = TRUE, sep = "\t", check.names = FALSE)
head(S16_family_taxa_profile)
S16_samples <- colnames(S16_family_taxa_profile)

### Extract 16S samples with MG data
# Get MG sample names
MG_df_fam <- df_fam
MG_samples <- colnames(MG_df_fam)
MG_samples
MG_samples <- MG_samples[MG_samples != "Family"]
# Convert MG sample names to 16S-style sample IDs
MG_samples_16S_style <- MG_samples %>%
  gsub("_all$", "", .) %>%   # remove trailing "_all"
  gsub("_", "-", .)          # replace "_" with "-"

MG_samples_16S_style

### Select 16S samples that having MG-based profilings
selected_16S_cols <- c("Genus", intersect(S16_samples, MG_samples_16S_style))
selected_16S_family_taxa_profile <- S16_family_taxa_profile[, selected_16S_cols]
# Check results
colnames(selected_16S_family_taxa_profile)
dim(selected_16S_family_taxa_profile)


top_n = 15
### Top 15 Families and Other Families in MG samples
# Convert into long table
MG_long_fam <- MG_df_fam %>%
  pivot_longer(
    cols = -Family,
    names_to = "Sample",
    values_to = "Abundance"
  ) %>%
  mutate(
    Abundance = as.numeric(Abundance),
    Abundance = replace_na(Abundance, 0)
  )

# Top families in MG data
MG_top_fam <- MG_long_fam %>%
  group_by(Family) %>%
  summarise(total = sum(Abundance, na.rm = TRUE), .groups = "drop") %>%
  arrange(desc(total)) %>%
  slice_head(n = top_n) %>%
  pull(Family)

# Combine left families into Other taxa
MG_plot_df <- MG_long_fam  %>%
  mutate(Family_plot = if_else(Family %in% MG_top_fam, Family, "Other")) %>%
  group_by(Sample, Family_plot) %>%
  summarise(Abundance = sum(Abundance, na.rm = TRUE), .groups = "drop")

head(MG_plot_df)
# Convert MG sample names to 16S-style sample IDs
MG_plot_df <- MG_plot_df %>%
  mutate(
    Sample = Sample %>%
      gsub("_all$", "", .) %>%
      gsub("_", "-", .)
  )

head(MG_plot_df)
tail(MG_plot_df)




### Top 15 Families and Other Families in 16S samples
head(selected_16S_family_taxa_profile)

# Convert into long table
S16_long_fam <- selected_16S_family_taxa_profile %>%
  pivot_longer(
    cols = -Genus,
    names_to = "Sample",
    values_to = "Abundance"
  ) %>%
  mutate(
    Abundance = as.numeric(Abundance),
    Abundance = replace_na(Abundance, 0)
  )

# Calculate family relative abundance in the samples
S16_long_fam <- S16_long_fam %>%
  group_by(Sample) %>%
  mutate(Rel_Abundance = Abundance / sum(Abundance, na.rm = TRUE)) %>%
  ungroup()

# Top families in 16S data
S16_top_fam <- S16_long_fam %>%
  group_by(Genus) %>%
  summarise(total = sum(Rel_Abundance, na.rm = TRUE), .groups = "drop") %>%
  arrange(desc(total)) %>%
  slice_head(n = top_n) %>%
  pull(Genus)
S16_top_fam

# Combine left families into Other taxa
S16_plot_df <- S16_long_fam  %>%
  mutate(Family_plot = if_else(Genus %in% S16_top_fam, Genus, "Other")) %>%
  group_by(Sample, Family_plot) %>%
  summarise(Abundance = sum(Abundance, na.rm = TRUE), Rel_Abundance = sum(Rel_Abundance, na.rm = TRUE), .groups = "drop")

head(S16_plot_df)


### Get the name of Top Families in both MG and 16S
top_genus_union <- union(MG_top_fam, S16_top_fam)


###### Regenerate 16S and MG tables
### MG tables
# Convert into long table
update_MG_long_fam <- MG_df_fam %>%
  pivot_longer(
    cols = -Family,
    names_to = "Sample",
    values_to = "Abundance"
  ) %>%
  mutate(
    Abundance = as.numeric(Abundance),
    Abundance = replace_na(Abundance, 0)
  )
# Combine left families into Other taxa
update_MG_plot_df <- update_MG_long_fam  %>%
  mutate(Family_plot = if_else(Family %in% top_genus_union, Family, "Other")) %>%
  group_by(Sample, Family_plot) %>%
  summarise(Abundance = sum(Abundance, na.rm = TRUE), .groups = "drop")

### S16 tables
# Convert into long table
update_S16_long_fam <- selected_16S_family_taxa_profile %>%
  pivot_longer(
    cols = -Genus,
    names_to = "Sample",
    values_to = "Abundance"
  ) %>%
  mutate(
    Abundance = as.numeric(Abundance),
    Abundance = replace_na(Abundance, 0)
  )

update_S16_long_fam <- update_S16_long_fam %>%
  group_by(Sample) %>%
  mutate(Rel_Abundance = Abundance / sum(Abundance, na.rm = TRUE)) %>%
  ungroup()

# Combine left families into Other taxa
update_S16_plot_df <- update_S16_long_fam  %>%
  mutate(Family_plot = if_else(Genus %in% top_genus_union, Genus, "Other")) %>%
  group_by(Sample, Family_plot) %>%
  summarise(
    Abundance = sum(Abundance, na.rm = TRUE),
    Rel_Abundance = sum(Rel_Abundance, na.rm = TRUE),
    .groups = "drop"
  )

head(update_S16_plot_df)
### Calculate relative abundance


###### Barplots
### Using a same color set for both 16S data and MG data
all_taxa <- unique(c(sort(unique(top_genus_union)), "Other"))
# pal <- colorRampPalette(brewer.pal(12, "Set3"))(length(all_taxa))
# names(pal) <- all_taxa
# # Reorder palette to match all_taxa order
# pal <- pal[all_taxa]
# names(pal) <- all_taxa
# pal["Other"] <- "grey70"   # force grey (safe even if already grey)

taxa_no_other <- setdiff(all_taxa, "Other")

base_cols <- c(
  brewer.pal(8, "Set2"),
  brewer.pal(12, "Set3"),
  brewer.pal(8, "Dark2"),
  brewer.pal(9, "Set1")
)

bad_cols <- c(
  "#D9D9D9", # grey, too close to Other
  "#B3B3B3", # grey, too close to Other
  "#FFFFB3", # very pale yellow
  "#B3DE69", # green, too close to other greens
  "#FFED6F"  # yellow, sometimes too close to other yellows
)


base_cols <- base_cols[!toupper(base_cols) %in% toupper(bad_cols)]

pal <- base_cols[seq_along(taxa_no_other)]
names(pal) <- taxa_no_other

pal <- c(pal, Other = "grey70")
pal <- pal[all_taxa]


# Fix Taxon order (optional but recommended)
update_MG_plot_df <- update_MG_plot_df %>%
  mutate(Family_plot = factor(Family_plot, levels = all_taxa))
update_S16_plot_df <- update_S16_plot_df %>%
  mutate(Family_plot = factor(Family_plot, levels = all_taxa))


### Plot (MG)
update_MG_plot_df <- update_MG_plot_df %>% mutate(Sample = Sample %>%
      gsub("_all$", "", .) %>%
      gsub("_", "-", .))

p_genus_MG <- ggplot(update_MG_plot_df, aes(x = Sample, y = Abundance, fill = Family_plot)) +
  geom_col(width = 0.9) +
  scale_fill_manual(values = pal) +
  labs(x = "Sample", y = "Relative abundance (%)", fill = "Genus") +
  theme_bw(base_size = 12) +
  theme(
    axis.text.x = element_text(angle = 75, hjust = 1, vjust = 1),
    panel.grid.major.x = element_blank()
  )
print(p_genus_MG)

### Saving MG Genus Figures
# figure_file <- file.path("../../tmp_figures/Taxonomics_Plot", "recolored_MG_based_genus_level_tax_profile.jpg")
# ggsave(figure_file, p_genus_MG, width = 12, height = 6, dpi = 300)
# figure_file <- file.path("../../tmp_figures/Taxonomics_Plot", "recolored_MG_based_genus_level_tax_profile.pdf")
# ggsave(figure_file, p_genus_MG, width = 12, height = 6)
# figure_file <- file.path("../../tmp_figures/Taxonomics_Plot", "recolored_MG_based_genus_level_tax_profile.svg")
# ggsave(figure_file,p_genus_MG, width = 12, height = 6)


### Plot (16S)
p_genus_S16 <- ggplot(update_S16_plot_df, aes(x = Sample, y = Rel_Abundance*100, fill = Family_plot)) +
  geom_col(width = 0.9) +
  scale_fill_manual(values = pal) +
  labs(x = "Sample", y = "Relative abundance (%)", fill = "Genus") +
  theme_bw(base_size = 12) +
  theme(
    axis.text.x = element_text(angle = 75, hjust = 1, vjust = 1),
    panel.grid.major.x = element_blank()
  )
print(p_genus_S16)

### Saving S16 Genus Figures
# figure_file <- file.path("../../tmp_figures/Taxonomics_Plot", "recolored_S16_based_genus_level_tax_profile.jpg")
# ggsave(figure_file, p_genus_S16, width = 12, height = 6, dpi = 300)
# figure_file <- file.path("../../tmp_figures/Taxonomics_Plot", "recolored_S16_based_genus_level_tax_profile.pdf")
# ggsave(figure_file, p_genus_S16, width = 12, height = 6)
# figure_file <- file.path("../../tmp_figures/Taxonomics_Plot", "recolored_S16_based_genus_level_tax_profile.svg")
# ggsave(figure_file, p_genus_S16, width = 12, height = 6)


# Saving family level plot dataframe
update_MG_genus_plot_df <- update_MG_plot_df
update_S16_genus_plot_df <- update_S16_plot_df
genus_pal <- pal

# ###### 16S and MG Grouped Plots
# library(dplyr)
# library(ggplot2)
# 
# ### Make sure MG sample names match 16S style
# update_MG_plot_df <- update_MG_plot_df %>%
#   mutate(
#     Sample = Sample %>%
#       gsub("_all$", "", .) %>%
#       gsub("_", "-", .)
#   )
# 
# ### If MG already has relative abundance in percent, rename it clearly;
# ### otherwise replace MG_abundance_pct with the correct relative abundance column
# MG_plot_ready <- update_MG_plot_df %>%
#   mutate(
#     Method = "Metagenomics",
#     Abundance_pct = Abundance   # 推荐这样；如果你现在只有百分比原值，再相应替换
#   ) %>%
#   select(Sample, Family_plot, Method, Abundance_pct)
# 
# S16_plot_ready <- update_S16_plot_df %>%
#   mutate(
#     Method = "16S rRNA",
#     Abundance_pct = Rel_Abundance * 100
#   ) %>%
#   select(Sample, Family_plot, Method, Abundance_pct)
# 
# ### Combine two tables
# plot_df <- bind_rows(MG_plot_ready, S16_plot_ready)
# 
# ### Fix sample order
# sample_order <- sort(unique(plot_df$Sample))
# plot_df <- plot_df %>%
#   mutate(
#     Sample = factor(Sample, levels = sample_order),
#     Method = factor(Method, levels = c("16S rRNA", "Metagenomics")),
#     Family_plot = factor(Family_plot, levels = all_taxa)
#   )
# 
# ### Combined plot
# p_combined <- ggplot(plot_df, aes(x = Sample, y = Abundance_pct, fill = Family_plot)) +
#   geom_col(width = 0.9) +
#   facet_wrap(~ Method, ncol = 1) +
#   scale_fill_manual(values = pal, guide = guide_legend(ncol = 1)) +
#   labs(x = "Sample", y = "Relative abundance (%)", fill = "Genus") +
#   theme_bw(base_size = 12) +
#   theme(
#     axis.text.x = element_text(angle = 75, hjust = 1, vjust = 1),
#     panel.grid.major.x = element_blank(),
#     strip.text = element_text(face = "bold"),
#     legend.position = "right"
#   )
# 
# print(p_combined)

# figure_file <- file.path("../../tmp_figures/", "recolored_S16_MG_genus_level_tax_profile.jpg")
# ggsave(figure_file, p_combined, width = 12, height = 8, dpi = 300)
```



### Combine 16S and MG taxonomic results at both Family and Genus Level
```{r}
library(patchwork)
library(ggtext)
# Get the following figures from previous two sections
# p_fam_MG, p_fam_S16, p_genus_MG, p_genus_S16

# Get the following plot dataframe from previous two sections
#update_MG_family_plot_df, update_S16_family_plot_df
#update_MG_genus_plot_df, update_S16_genus_plot_df

# Get the following color assignment from previous two sections
# fam_pal, genus_pal

# Set Figure Theme
pub_theme <- theme_bw(base_size = 12) +
  theme(
    panel.grid.major.x = element_blank(),
    panel.grid.minor = element_blank(),
    plot.title = element_text(face = "bold", size = 14),
    axis.title = element_text(face = "bold"),
    legend.title = element_text(face = "bold"),
    legend.position = "right",
    legend.box = "vertical",
    legend.text = element_text(size = 9)
  )

top_theme <- pub_theme +
  theme(
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank(),
    axis.title.x = element_blank()
  )

bottom_theme <- pub_theme +
  theme(
    axis.text.x = element_text(angle = 75, hjust = 1, vjust = 1)
  )


# Update 16S Genus Names
final_update_S16_genus_plot_df <- update_S16_genus_plot_df %>%
  mutate(
    Family_plot = recode(
      Family_plot,
      "[Ruminococcus]_torques_group" = "[Ruminococcus] torques\ngroup",
      "Tannerellaceae" = "Unclassified Tannerellaceae"
    )
  )

S16_genus_pal <- genus_pal
names(S16_genus_pal)[names(S16_genus_pal) == "[Ruminococcus]_torques_group"] <- "[Ruminococcus] torques\ngroup"
names(S16_genus_pal)[names(S16_genus_pal) == "Tannerellaceae"] <- "Unclassified Tannerellaceae"

# Set text format of the 16S Genus Figure
S16_genus_labels <- c(
  "[Ruminococcus] torques\ngroup" = "[Ruminococcus] torques\ngroup",
  "Agathobacter" = "<i>Agathobacter</i>",
  "Akkermansia" = "<i>Akkermansia</i>",
  "Alistipes" = "<i>Alistipes</i>",
  "Bacteroides" = "<i>Bacteroides</i>",
  "Bifidobacterium" = "<i>Bifidobacterium</i>",
  "Blautia" = "<i>Blautia</i>",
  "Collinsella" = "<i>Collinsella</i>",
  "Escherichia-Shigella" = "<i>Escherichia</i>-<i>Shigella</i>",
  "Faecalibacterium" = "<i>Faecalibacterium</i>",
  "Odoribacter" = "<i>Odoribacter</i>",
  "Parabacteroides" = "<i>Parabacteroides</i>",
  "Roseburia" = "<i>Roseburia</i>",
  "Ruminococcus" = "<i>Ruminococcus</i>",
  "Streptococcus" = "<i>Streptococcus</i>",
  "Unclassified Tannerellaceae" = "Unclassified Tannerellaceae",
  "Unknown" = "Unknown",
  "Other" = "Other"
)


# Update 16S Genus Names
final_update_MG_genus_plot_df <- update_MG_genus_plot_df %>%
  mutate(
    Family_plot = recode(
      Family_plot,
      "Lachnospiraceae_unclassified" = "Unclassified Lachnospiraceae"
    )
  )

MG_genus_pal <- genus_pal
names(MG_genus_pal)[names(MG_genus_pal) == "Lachnospiraceae_unclassified"] <- "Unclassified Lachnospiraceae"


# Set text format of the MG Genus Figure
MG_genus_labels <- c(
  "Akkermansia" = "<i>Akkermansia</i>",
  "Alistipes" = "<i>Alistipes</i>",
  "Bacteroides" = "<i>Bacteroides</i>",
  "Bifidobacterium" = "<i>Bifidobacterium</i>",
  "Blautia" = "<i>Blautia</i>",
  "Clostridium" = "<i>Clostridium</i>",
  "Collinsella" = "<i>Collinsella</i>",
  "Escherichia" = "<i>Escherichia</i>",
  "Faecalibacterium" = "<i>Faecalibacterium</i>",
  "Unclassified Lachnospiraceae" = "Unclassified Lachnospiraceae",
  "Mediterraneibacter" = "<i>Mediterraneibacter</i>",
  "Odoribacter" = "<i>Odoribacter</i>",
  "Parabacteroides" = "<i>Parabacteroides</i>",
  "Phocaeicola" = "<i>Phocaeicola</i>",
  "Roseburia" = "<i>Roseburia</i>",
  "Ruminococcus" = "<i>Ruminococcus</i>",
  "Streptococcus" = "<i>Streptococcus</i>",
  "Other" = "Other"
)


p_fam_MG <- ggplot(update_MG_fam_plot_df, aes(x = Sample, y = Abundance, fill = Family_plot)) +
  geom_col(width = 0.9) +
  scale_fill_manual(values = fam_pal) +
  # labs(title = "MG-based family profile",
  #      x = NULL, y = "Relative abundance (%)", fill = "Family") +
  labs(x = NULL, y = "Relative abundance (%)", fill = "Family") +
  top_theme
p_fam_MG 

p_fam_S16 <- ggplot(update_S16_fam_plot_df, aes(x = Sample, y = Rel_Abundance * 100, fill = Family_plot)) +
  geom_col(width = 0.9) +
  scale_fill_manual(values = fam_pal) +
  # labs(title = "16S-based family profile",
  #      x = NULL, y = "Relative abundance (%)", fill = "Family") +
  labs(x = NULL, y = "Relative abundance (%)", fill = "Family") +
  top_theme
p_fam_S16

p_genus_MG <- ggplot(final_update_MG_genus_plot_df, aes(x = Sample, y = Abundance, fill = Family_plot)) +
  geom_col(width = 0.9) +
  scale_fill_manual(values = MG_genus_pal, labels = MG_genus_labels) +
  # labs(title = "MG-based genus profile",
  #      x = "Sample", y = "Relative abundance (%)", fill = "Genus") +
  labs(x = "Sample", y = "Relative abundance (%)", fill = "Genus") +  
  bottom_theme +
  theme(legend.text = ggtext::element_markdown())
p_genus_MG

p_genus_S16 <- ggplot(final_update_S16_genus_plot_df, aes(x = Sample, y = Rel_Abundance * 100, fill = Family_plot)) +
  geom_col(width = 0.9) +
  scale_fill_manual(values = S16_genus_pal, labels = S16_genus_labels) +
  # labs(title = "16S-based genus profile",
  #      x = "Sample", y = "Relative abundance (%)", fill = "Genus") +
  labs(x = "Sample", y = "Relative abundance (%)", fill = "Genus") +
  bottom_theme + 
  theme(legend.text = ggtext::element_markdown())
p_genus_S16

combined_S16 <- p_fam_S16 / p_genus_S16 +
  plot_layout(heights = c(1, 1))
combined_S16

### Saving combined S16 Plots
figure_file <- file.path("../../results/MG_results/Taxonomics_results", "S16_based_both_level_tax_profile.jpg")
ggsave(figure_file, combined_S16, width = 12, height = 12, dpi = 300)
# figure_file <- file.path("../../results/MG_results/Taxonomics_results", "S16_based_both_level_tax_profile.pdf")
# ggsave(figure_file, combined_S16, width = 12, height = 12)
# figure_file <- file.path("../../results/MG_results/Taxonomics_results", "S16_based_both_level_tax_profile.svg")
# ggsave(figure_file,combined_S16, width = 12, height = 12)


combined_MG <- p_fam_MG / p_genus_MG +
  plot_layout(heights = c(1, 1))
combined_MG

### Saving combined MG Plots
figure_file <- file.path("../../results/MG_results/Taxonomics_results", "MG_based_both_level_tax_profile.jpg")
ggsave(figure_file, combined_MG, width = 12, height = 12, dpi = 300)
# figure_file <- file.path("../../results/MG_results/Taxonomics_results", "MG_based_both_level_tax_profile.pdf")
# ggsave(figure_file, combined_MG, width = 12, height = 12)
# figure_file <- file.path("../../results/MG_results/Taxonomics_results", "MG_based_both_level_tax_profile.svg")
# ggsave(figure_file,combined_MG, width = 12, height = 12)
```



### 16S Taxonomic Profiling for All Samples (Family Level)
```{r}
library(ggplot2)

###### Read in 16S-based taxonomic profilings
S16_family_taxa_file <- file.path("../../data/Metagenomics/16S_related", "all_samples_16S_Family_Profiling.tsv")

S16_family_taxa_profile <- read.table(S16_family_taxa_file, header = TRUE, sep = "\t", check.names = FALSE)
head(S16_family_taxa_profile)

colnames(S16_family_taxa_profile)
head(S16_family_taxa_profile)


### Assign Top Taxa (From the Previous Family Section)
top_fam_union

### Taxa Color Assignment (From the Previous Family Section)
fam_pal

# Convert into long table
all_S16_long_fam <- S16_family_taxa_profile %>%
  pivot_longer(
    cols = -Family,
    names_to = "Sample",
    values_to = "Abundance"
  ) %>%
  mutate(
    Abundance = as.numeric(Abundance),
    Abundance = replace_na(Abundance, 0)
  )

# Calculate family relative abundance in the samples
all_S16_long_fam <- all_S16_long_fam %>%
  group_by(Sample) %>%
  mutate(Rel_Abundance = Abundance / sum(Abundance, na.rm = TRUE)) %>%
  ungroup()


# Combine left families into Other taxa
all_S16_plot_df <- all_S16_long_fam  %>%
  mutate(Family_plot = if_else(Family %in% top_fam_union, Family, "Other")) %>%
  group_by(Sample, Family_plot) %>%
  summarise(Abundance = sum(Abundance, na.rm = TRUE), Rel_Abundance = sum(Rel_Abundance, na.rm = TRUE), .groups = "drop")

all_S16_plot_df <- all_S16_plot_df %>%
  mutate(Individual = str_extract(Sample, "^[^-]+"))

head(all_S16_plot_df)
tail(all_S16_plot_df)


### Update Family Names
final_all_S16_plot_df <- all_S16_plot_df %>%
  mutate(
    Family_plot = recode(
      Family_plot,
      "Eubacteriales_unclassified" = "Unclassified Eubacteriales"
    )
  )

final_fam_pal <- fam_pal
names(final_fam_pal)[names(final_fam_pal) == "Eubacteriales_unclassified"] <- "Unclassified Eubacteriales"

### Set Family Orders
fam_order <- c(
  setdiff(names(final_fam_pal), "Other"),
  "Other"
)

final_all_S16_plot_df <- final_all_S16_plot_df %>%
  mutate(
    Family_plot = factor(Family_plot, levels = fam_order)
  )





# ###### Save into Six Different Figures
n_per_page <- 8
out_dir <- "../../results/MG_results/Taxonomics_results"
dir.create(out_dir, showWarnings = FALSE, recursive = TRUE)

# # 1)（可选）让 Sample 在每个 Individual 内按编号排序，避免 NW001-10 排在 NW001-2 前面
plot_df <- final_all_S16_plot_df %>%
  mutate(
    Sample_num = as.integer(str_extract(Sample, "(?<=-)\\d+$"))
  ) %>%
  arrange(Individual, Sample_num) %>%
  mutate(Sample = factor(Sample, levels = unique(Sample)))
#
# # 2) 给 Individual 分页编号：每 n_per_page 个为一组
ind_levels <- sort(unique(plot_df$Individual))
ind_page_map <- tibble(
  Individual = ind_levels,
  Page = ceiling(seq_along(ind_levels) / n_per_page)
)

plot_df <- plot_df %>%
  left_join(ind_page_map, by = "Individual")

n_pages <- max(plot_df$Page)

# # 3) 循环画图 + 保存
for (pg in seq_len(n_pages)) {
  df_pg <- plot_df %>% filter(Page == pg)

  p <- ggplot(df_pg, aes(x = Sample, y = Rel_Abundance*100, fill = Family_plot)) +
    geom_col(width = 0.9) +
    facet_wrap(~ Individual, ncol = 4, scales = "free_x") +  # ncol可调，比如 2/3/4
    scale_fill_manual(values = final_fam_pal) +
    labs(
      title = paste0("Taxonomic profiling (Individuals ", (pg-1)*n_per_page + 1, "–", min(pg*n_per_page, length(ind_levels)), ")"),
      x = "Sample", y = "Relative abundance (%)", fill = "Family"
    ) +
    theme_bw(base_size = 12) +
    theme(
      axis.text.x = element_text(angle = 75, hjust = 1, vjust = 1),
      panel.grid.major.x = element_blank(),
      strip.text = element_text(face = "bold"),
      legend.position = "right"
    )

  print(p)

  # ggsave(
  #   filename = file.path(out_dir, sprintf("S16_tax_profile_page_%02d.jpg", pg)),
  #   plot = p, width = 14, height = 8, dpi = 300
  # )
}



###### Save into One Large Figure
plot_df <- final_all_S16_plot_df %>%
  mutate(
    Sample_num1 = as.integer(str_extract(Sample, "(?<=-)\\d+")),
    Sample_num2 = as.integer(str_extract(Sample, "(?<=-\\d-)\\d+$")),
    Sample_num2 = ifelse(is.na(Sample_num2), 0, Sample_num2)
  ) %>%
  arrange(Individual, Sample_num1, Sample_num2) %>%
  mutate(Sample = factor(Sample, levels = unique(Sample)))

n_per_page <- 8

ind_levels <- sort(unique(plot_df$Individual))

ind_page_map <- tibble(
  Individual = ind_levels,
  Page = ceiling(seq_along(ind_levels) / n_per_page)
)

plot_df <- plot_df %>%
  left_join(ind_page_map, by = "Individual")

n_pages <- max(plot_df$Page)

pub_theme <- theme_bw(base_size = 12) +
  theme(
    axis.text.x = element_text(angle = 75, hjust = 1, vjust = 1, size = 11),
    axis.text.y = element_text(size = 11),
    axis.title = element_text(face = "bold", size = 13),
    panel.grid.major.x = element_blank(),
    panel.grid.minor = element_blank(),
    strip.text = element_text(face = "bold", size = 13),
    plot.title = element_text(face = "bold", size = 14, hjust = 0.5),
    legend.title = element_text(face = "bold", size = 16),
    legend.text = element_text(size = 14)
  )


legend_plot <- ggplot(plot_df, aes(x = Sample, y = Rel_Abundance * 100, fill = Family_plot)) +
  geom_col(width = 0.9) +
  facet_wrap(~ Individual, ncol = 3, scales = "free_x") +
  scale_fill_manual(values = final_fam_pal) +
  labs(x = "Sample", y = "Relative abundance (%)", fill = "Family") +
  pub_theme +
  theme(legend.position = "right")

legend_grob <- cowplot::get_legend(legend_plot)

plot_list <- vector("list", n_pages)

for (pg in seq_len(n_pages)) {
  df_pg <- plot_df %>% filter(Page == pg)

  p <- ggplot(df_pg, aes(x = Sample, y = Rel_Abundance * 100, fill = Family_plot)) +
    geom_col(width = 0.9) +
    facet_wrap(~ Individual, ncol = 4, scales = "free_x") +
    scale_fill_manual(values = final_fam_pal) +
    labs(
      x = "Sample",
      y = "Relative abundance (%)",
      fill = "Family"
    ) +
    pub_theme +
    theme(legend.position = "none")

  plot_list[[pg]] <- p
}

combined_panels <- cowplot::plot_grid(
  plotlist = plot_list,
  ncol = 2,
  align = "hv"
)

final_combined_plot <- cowplot::plot_grid(
  combined_panels,
  legend_grob,
  ncol = 2,
  rel_widths = c(4.5, 1.2)
)

final_combined_plot



out_dir <- "../../results/MG_results/Taxonomics_results"
dir.create(out_dir, showWarnings = FALSE, recursive = TRUE)

ggsave(
  filename = file.path(out_dir, "All_S16_tax_profile_combined.jpg"),
  plot = final_combined_plot,
  width = 16,
  height = 20,
  dpi = 300
)

# ggsave(
#   filename = file.path(out_dir, "All_S16_tax_profile_combined.pdf"),
#   plot = final_combined_plot,
#   width = 16,
#   height = 20,
# )
# 
# ggsave(
#   filename = file.path(out_dir, "All_S16_tax_profile_combined.svg"),
#   plot = final_combined_plot,
#   width = 16,
#   height = 20,
# )

```



### 16S Taxonomic Profiling for All Samples (Genus Level, not shown in the manuscripts)
```{r}
library(ggplot2)
library(ggtext)

###### Read in 16S-based taxonomic profilings
S16_family_taxa_file <- file.path("../../data/Metagenomics/16S_related", "all_samples_16S_Genus_Profiling.tsv")

S16_family_taxa_profile <- read.table(S16_family_taxa_file, header = TRUE, sep = "\t", check.names = FALSE)
head(S16_family_taxa_profile)

colnames(S16_family_taxa_profile)
head(S16_family_taxa_profile)


### Assign Top Taxa (From the Previous Genus Section)
top_fam_union <- top_genus_union

### Taxa Color Assignment (From the Previous Family Section)
fam_pal <- genus_pal

# Convert into long table
all_S16_long_fam <- S16_family_taxa_profile %>%
  pivot_longer(
    cols = -Genus,
    names_to = "Sample",
    values_to = "Abundance"
  ) %>%
  mutate(
    Abundance = as.numeric(Abundance),
    Abundance = replace_na(Abundance, 0)
  )

# Calculate family relative abundance in the samples
all_S16_long_fam <- all_S16_long_fam %>%
  group_by(Sample) %>%
  mutate(Rel_Abundance = Abundance / sum(Abundance, na.rm = TRUE)) %>%
  ungroup()


# Combine left families into Other taxa
all_S16_plot_df <- all_S16_long_fam  %>%
  mutate(Family_plot = if_else(Genus %in% top_fam_union, Genus, "Other")) %>%
  group_by(Sample, Family_plot) %>%
  summarise(Abundance = sum(Abundance, na.rm = TRUE), Rel_Abundance = sum(Rel_Abundance, na.rm = TRUE), .groups = "drop")

all_S16_plot_df <- all_S16_plot_df %>%
  mutate(Individual = str_extract(Sample, "^[^-]+"))

head(all_S16_plot_df)
tail(all_S16_plot_df)



unique(all_S16_plot_df$Family_plot)

### Update Genus Names
final_all_S16_plot_df <- all_S16_plot_df %>%
  mutate(
    Family_plot = recode(
      Family_plot,
      "[Ruminococcus]_torques_group" = "[Ruminococcus] torques group",
      "Tannerellaceae" = "Unclassified Tannerellaceae"
    )
  )


final_genus_pal <- genus_pal
names(final_genus_pal)[names(final_genus_pal) == "[Ruminococcus]_torques_group"] <- "[Ruminococcus] torques group"

names(final_genus_pal)[names(final_genus_pal) == "Tannerellaceae"] <- "Unclassified Tannerellaceae"


### Set Family Orders
fam_order <- c(
  setdiff(names(final_genus_pal), "Other"),
  "Other"
)

final_all_S16_plot_df <- final_all_S16_plot_df %>%
  mutate(
    Family_plot = factor(Family_plot, levels = fam_order)
  )





# ###### Save into Six Different Figures
n_per_page <- 8   # 改成 8 或 10
out_dir <- "../../results/MG_results/Taxonomics_results"
dir.create(out_dir, showWarnings = FALSE, recursive = TRUE)

# # 1)（可选）让 Sample 在每个 Individual 内按编号排序，避免 NW001-10 排在 NW001-2 前面
plot_df <- final_all_S16_plot_df %>%
  mutate(
    Sample_num1 = as.integer(str_extract(Sample, "(?<=-)\\d+")),
    Sample_num2 = as.integer(str_extract(Sample, "(?<=-\\d-)\\d+$")),
    Sample_num2 = ifelse(is.na(Sample_num2), 0, Sample_num2)
  ) %>%
  arrange(Individual, Sample_num1, Sample_num2) %>%
  mutate(Sample = factor(Sample, levels = unique(Sample)))

#
# # 2) 给 Individual 分页编号：每 n_per_page 个为一组
ind_levels <- sort(unique(plot_df$Individual))
ind_page_map <- tibble(
  Individual = ind_levels,
  Page = ceiling(seq_along(ind_levels) / n_per_page)
)

plot_df <- plot_df %>%
  left_join(ind_page_map, by = "Individual")

n_pages <- max(plot_df$Page)


# Set text format of the Genus Figure
S16_genus_labels <- c(
  "[Ruminococcus] torques group" = "[Ruminococcus] torques group",
  "Agathobacter" = "<i>Agathobacter</i>",
  "Akkermansia" = "<i>Akkermansia</i>",
  "Alistipes" = "<i>Alistipes</i>",
  "Bacteroides" = "<i>Bacteroides</i>",
  "Bifidobacterium" = "<i>Bifidobacterium</i>",
  "Blautia" = "<i>Blautia</i>",
  "Collinsella" = "<i>Collinsella</i>",
  "Escherichia-Shigella" = "<i>Escherichia</i>-<i>Shigella</i>",
  "Faecalibacterium" = "<i>Faecalibacterium</i>",
  "Odoribacter" = "<i>Odoribacter</i>",
  "Parabacteroides" = "<i>Parabacteroides</i>",
  "Roseburia" = "<i>Roseburia</i>",
  "Ruminococcus" = "<i>Ruminococcus</i>",
  "Streptococcus" = "<i>Streptococcus</i>",
  "Unclassified Tannerellaceae" = "Unclassified Tannerellaceae",
  "Unknown" = "Unknown",
  "Other" = "Other"
)

# # 3) 循环画图 + 保存
for (pg in seq_len(n_pages)) {
  df_pg <- plot_df %>% filter(Page == pg)

  p <- ggplot(df_pg, aes(x = Sample, y = Rel_Abundance*100, fill = Family_plot)) +
    geom_col(width = 0.9) +
    facet_wrap(~ Individual, ncol = 4, scales = "free_x") +  # ncol可调，比如 2/3/4
    scale_fill_manual(values = final_genus_pal, labels = S16_genus_labels) +
    labs(
      title = paste0("Taxonomic profiling (Individuals ", (pg-1)*n_per_page + 1, "–", min(pg*n_per_page, length(ind_levels)), ")"),
      x = "Sample", y = "Relative abundance (%)", fill = "Genus"
    ) +
    theme_bw(base_size = 12) +
    theme(
      axis.text.x = element_text(angle = 75, hjust = 1, vjust = 1),
      panel.grid.major.x = element_blank(),
      strip.text = element_text(face = "bold"),
      legend.position = "right",
      legend.text = ggtext::element_markdown()
    )

  print(p)

  # ggsave(
  #   filename = file.path(out_dir, sprintf("S16_tax_genus_profile_page_%02d.jpg", pg)),
  #   plot = p, width = 14, height = 8, dpi = 300
  # )
}



###### Save into One Large Figure
plot_df <- final_all_S16_plot_df %>%
  mutate(
    Sample_num1 = as.integer(str_extract(Sample, "(?<=-)\\d+")),
    Sample_num2 = as.integer(str_extract(Sample, "(?<=-\\d-)\\d+$")),
    Sample_num2 = ifelse(is.na(Sample_num2), 0, Sample_num2)
  ) %>%
  arrange(Individual, Sample_num1, Sample_num2) %>%
  mutate(Sample = factor(Sample, levels = unique(Sample)))

n_per_page <- 8

ind_levels <- sort(unique(plot_df$Individual))

ind_page_map <- tibble(
  Individual = ind_levels,
  Page = ceiling(seq_along(ind_levels) / n_per_page)
)

plot_df <- plot_df %>%
  left_join(ind_page_map, by = "Individual")

n_pages <- max(plot_df$Page)

pub_theme <- theme_bw(base_size = 12) +
  theme(
    axis.text.x = element_text(angle = 75, hjust = 1, vjust = 1, size = 11),
    axis.text.y = element_text(size = 11),
    axis.title = element_text(face = "bold", size = 13),
    panel.grid.major.x = element_blank(),
    panel.grid.minor = element_blank(),
    strip.text = element_text(face = "bold", size = 13),
    plot.title = element_text(face = "bold", size = 14, hjust = 0.5),
    legend.title = element_text(face = "bold", size = 16),
    legend.text = element_text(size = 14)
  )


legend_plot <- ggplot(plot_df, aes(x = Sample, y = Rel_Abundance * 100, fill = Family_plot)) +
  geom_col(width = 0.9) +
  facet_wrap(~ Individual, ncol = 3, scales = "free_x") +
  scale_fill_manual(values = final_genus_pal, labels = S16_genus_labels) +
  labs(x = "Sample", y = "Relative abundance (%)", fill = "Genus") +
  pub_theme +
  theme(legend.position = "right", legend.text = ggtext::element_markdown())

legend_grob <- cowplot::get_legend(legend_plot)

plot_list <- vector("list", n_pages)

for (pg in seq_len(n_pages)) {
  df_pg <- plot_df %>% filter(Page == pg)

  p <- ggplot(df_pg, aes(x = Sample, y = Rel_Abundance * 100, fill = Family_plot)) +
    geom_col(width = 0.9) +
    facet_wrap(~ Individual, ncol = 4, scales = "free_x") +
    scale_fill_manual(values = final_genus_pal) +
    labs(
      x = "Sample",
      y = "Relative abundance (%)",
      fill = "Family"
    ) +
    pub_theme +
    theme(legend.position = "none")

  plot_list[[pg]] <- p
}

combined_panels <- cowplot::plot_grid(
  plotlist = plot_list,
  ncol = 2,
  align = "hv"
)

final_combined_plot <- cowplot::plot_grid(
  combined_panels,
  legend_grob,
  ncol = 2,
  rel_widths = c(4.5, 1.2)
)

final_combined_plot



# Saving the figure
out_dir <- "../../results/MG_results/Taxonomics_results"
dir.create(out_dir, showWarnings = FALSE, recursive = TRUE)

ggsave(
  filename = file.path(out_dir, "All_S16_tax_genus_profile_combined.jpg"),
  plot = final_combined_plot,
  width = 16,
  height = 20,
  dpi = 300
)

# ggsave(
#   filename = file.path(out_dir, "All_S16_tax_genus_profile_combined.pdf"),
#   plot = final_combined_plot,
#   width = 16,
#   height = 20,
# )
# 
# ggsave(
#   filename = file.path(out_dir, "All_S16_tax_genus_profile_combined.svg"),
#   plot = final_combined_plot,
#   width = 16,
#   height = 20,
# )

```





### Aitchison and Bray-Curtis distance between metagenomics and 16S-based family level taxonomic profiling
```{r}
#16S taxonomic profile from previous section (Family level)
S16_plot_df

#MG taxonomic profile from previous section (Family level)
MG_plot_df



library(tidyverse)
library(vegan)
library(compositions)
library(ggplot2)
library(cowplot)
library(tibble)

### Step 1. Prepare relative abundance for MG
MG_plot_df2 <- MG_plot_df %>%
  group_by(Sample) %>%
  mutate(Rel_Abundance = Abundance / sum(Abundance, na.rm = TRUE)) %>%
  ungroup()

### Step 2. Keep matched samples
matched_samples <- intersect(unique(S16_plot_df$Sample), unique(MG_plot_df2$Sample))

S16_sub <- S16_plot_df %>% filter(Sample %in% matched_samples)
MG_sub  <- MG_plot_df2 %>% filter(Sample %in% matched_samples)

### Step 3. Union of families
all_families <- union(unique(S16_sub$Family_plot), unique(MG_sub$Family_plot))

### Step 4. Wide matrices
S16_mat <- S16_sub %>%
  select(Sample, Family_plot, Rel_Abundance) %>%
  complete(Sample = matched_samples, Family_plot = all_families, fill = list(Rel_Abundance = 0)) %>%
  pivot_wider(names_from = Family_plot, values_from = Rel_Abundance) %>%
  arrange(Sample)

MG_mat <- MG_sub %>%
  select(Sample, Family_plot, Rel_Abundance) %>%
  complete(Sample = matched_samples, Family_plot = all_families, fill = list(Rel_Abundance = 0)) %>%
  pivot_wider(names_from = Family_plot, values_from = Rel_Abundance) %>%
  arrange(Sample)

stopifnot(all(S16_mat$Sample == MG_mat$Sample))

S16_mat2 <- S16_mat %>%
  column_to_rownames("Sample") %>%
  as.matrix()

MG_mat2 <- MG_mat %>%
  column_to_rownames("Sample") %>%
  as.matrix()

### Step 5. Distance functions
calc_aitchison <- function(x, y, pseudocount = 1e-6) {
  x <- x + pseudocount
  y <- y + pseudocount
  x <- x / sum(x)
  y <- y / sum(y)
  x_clr <- compositions::clr(x)
  y_clr <- compositions::clr(y)
  sqrt(sum((x_clr - y_clr)^2))
}

calc_bray <- function(x, y) {
  as.numeric(vegan::vegdist(rbind(x, y), method = "bray")[1])
}

### Step 6. Calculate distances
dist_df <- map_dfr(seq_len(nrow(S16_mat2)), function(i) {
  x <- S16_mat2[i, ]
  y <- MG_mat2[i, ]
  
  tibble(
    Sample = rownames(S16_mat2)[i],
    Bray_Curtis = calc_bray(x, y),
    Aitchison = calc_aitchison(x, y, pseudocount = 1e-6)
  )
})

print(dist_df)

### Step 7. Summary stats
aitchison_mean <- mean(dist_df$Aitchison, na.rm = TRUE)
aitchison_se <- sd(dist_df$Aitchison, na.rm = TRUE) / sqrt(sum(!is.na(dist_df$Aitchison)))

cat("Average Aitchison distance =", round(aitchison_mean, 4), "\n")
cat("SE =", round(aitchison_se, 4), "\n")

bc_mean <- mean(dist_df$Bray_Curtis, na.rm = TRUE)
bc_se <- sd(dist_df$Bray_Curtis, na.rm = TRUE) / sqrt(sum(!is.na(dist_df$Bray_Curtis)))

cat("Average Bray-Curtis distance =", round(bc_mean, 4), "\n")
cat("SE =", round(bc_se, 4), "\n")

### Step 8. Correlation
cor_res <- cor.test(dist_df$Bray_Curtis, dist_df$Aitchison, method = "spearman")

## correlation text
cor_label <- paste0(
  "Spearman's rho = ", round(cor_res$estimate, 3),
  "\nP = ", signif(cor_res$p.value, 3)
)

### Main scatter plot
p_scatter <- ggplot(dist_df, aes(x = Bray_Curtis, y = Aitchison)) +
  geom_point(size = 2.2, alpha = 0.8) +
  geom_smooth(method = "lm", se = FALSE, linewidth = 0.7, linetype = 2) +
  annotate(
    "text",
    x = Inf, y = 20,
    label = cor_label,
    hjust = 1.05, vjust = 1.3,
    size = 4
  ) +
  labs(
    x = "Bray-Curtis distance",
    y = "Aitchison distance"
  ) +
  theme_classic(base_size = 12) +
  theme(
    axis.title = element_text(size = 12),
    axis.text = element_text(size = 10),
    plot.margin = margin(5.5, 5.5, 5.5, 5.5)
  )

### Top marginal boxplot for Bray-Curtis
p_top <- ggplot(dist_df, aes(x = Bray_Curtis, y = "")) +
  geom_boxplot(width = 0.25, outlier.shape = NA) +
  geom_jitter(height = 0.08, width = 0, alpha = 0.7, size = 1.5) +
  labs(x = NULL, y = NULL) +
  theme_classic(base_size = 11) +
  theme(
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank(),
    axis.line.x = element_blank(),
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank(),
    axis.line.y = element_blank(),
    plot.margin = margin(5.5, 5.5, 0, 5.5)
  )

### Right marginal boxplot for Aitchison
p_right <- ggplot(dist_df, aes(x = "", y = Aitchison)) +
  geom_boxplot(width = 0.25, outlier.shape = NA) +
  geom_jitter(width = 0.08, height = 0, alpha = 0.7, size = 1.5) +
  labs(x = NULL, y = NULL) +
  theme_classic(base_size = 11) +
  theme(
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank(),
    axis.line.x = element_blank(),
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank(),
    axis.line.y = element_blank(),
    plot.margin = margin(5.5, 5.5, 5.5, 0)
  )

### Empty plot
p_empty <- ggplot() + theme_void()

### Combine
top_row <- plot_grid(
  p_top, p_empty,
  ncol = 2,
  rel_widths = c(4, 1),
  align = "h"
)

bottom_row <- plot_grid(
  p_scatter, p_right,
  ncol = 2,
  rel_widths = c(4, 1),
  align = "h"
)

combined_plot <- plot_grid(
  top_row, bottom_row,
  nrow = 2,
  rel_heights = c(1, 4),
  align = "v"
)

print(combined_plot)

### Step 10. Save the figure
figure_file <- file.path("../../results/MG_results/Taxonomics_results", "Family_level_Bray_vs_Aitchison_distance.jpg")
ggsave(figure_file, combined_plot, width = 8, height = 7, dpi = 300)
# write.table(dist_df, "Family_level_matched_sample_distances.txt", sep = "\t", quote = FALSE, row.names = FALSE)


```


Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
